### 2014-02-21 article.tex
template
git-svn-id: https://repos.deltares.nl/repos/phd/baart_f/documents/latex/articles/iemss2014@5383 acef1543-32c8-48d0-a996-87da9e3efd67

\documentclass[a4paper]{article}
\usepackage{iemss}
\usepackage[pdftex]{graphicx}
\usepackage{natbib}
\usepackage{amsmath}
\usepackage[labelfont=bf,textfont=sf,labelsep=period]{caption}
\usepackage[english]{babel}
%\RequirePackage{lineno} % for line numbers

\pdfpagewidth=210 true mm
\pdfpageheight=297 true mm

\renewcommand{\rmdefault}{phv} % Arial
\renewcommand{\sfdefault}{phv} % Arial

\bibpunct{[}{]}{;}{a}{,}{,~}

\pagestyle{IEMSSheadings}

\DeclareRobustCommand{\orderof}{\ensuremath{\mathcal{O}}}
\DeclareRobustCommand{\threedi}{3Di }

% Text to appear in the header of the pages
\IEMSShead{Baart et al. / Interactive web based flood modeling}

\graphicspath{{figures/}}
\DeclareGraphicsExtensions{.pdf,.png,.jpg}

\title{Interactive web based flood modeling at country wide scale and planter size resolution.}

%Authors Names and Affiliations:  Two spaces below the title, 10 pt
%Arial, Upper and Lower Case, underline author presenting
%paper and  provide his/her email address

\author{\underline{F. Baart}
\address[A1]{\it{Deltares,
Rotterdamseweg
Delft, The Netherlands (fedor.baart@deltares.nl, arthur.vandam@deltares.nl, gennadii.donchyts@deltares.nl)}},
J. Ha
\address[B1]{\it{Nelen \& Schuurmans,
Zakkendragerssteeg,
Utrecht, The Netherlands (jack.ha@nelen-schuurmans.nl, martijn.siemerink@nelen-schuurmans.nl)}},
A. van Dam \addressmark[A1],
G. Donchyts \addressmark[A1],
M. Siemerink \addressmark[B1]
}

\begin{document}

\begin{abstract}
The flooding of rural and urban areas is an increasing hazard to society. Accurate and timely predictions are essential for the water manager to prepare and respond to these hazards.
Predicting flooding requires a numerical model that represents the physical processes (rain, evaporation, infiltration, overland flow, groundwater flow). This model, fed with measurements, and possible measures, calculates the expected flooding.
The traditional modelling process consists of a three step process: schematization setup, model running and post-processing, with a total feedback time of hours.  This process is suitable for confirmatory modelling. Most of the time, models are applied exploratory, requiring a different workflow.
Enabling exploratory modelling requires a shift in utilisation of the modelling instrument. Stakeholders are in control and together evaluate ideas by interacting with the model through a mobile compatible website, supported by the modellers’ expertise. Enabling this type of interactive modelling requires a new level of performance.
The \threedi platform, in which the new approach was applied, consists of a new flooding and hydrological model (1D/2D) with a corresponding cloud based infrastructure. Applications in rural and urban areas of \orderof(1000km2) at a resolution of \orderof(0.1m) have shown its capabilities for both exploratory and confirmatory modelling.
The ambition that every component should be at least a 100 times faster than the previous approach, resulted in several advancements, both in the numerical engine and the software that interacts with the user and pushes the data to the web. Here we show advancements in the architecture and model communication.

\end{abstract}
\begin{keyword}
\end{keyword}

\maketitle

\section{Introduction}

The flooding of rural and urban areas is an increasing hazard to society. Accurate and timely predictions are essential for the water manager to be able to carefully plan, design and control the effects of flooding hazard.
For the prediction of these flooding hazards requires a numerical model that represents the physical processes (rain, evaporation, flow, infiltration). This model is fed with measurements from current or historic events. The results of these calculations, the expected flooding, is used for damage estimates, dike design, spatial and emergency planning and for coordinating response efforts.
Several challenges required an innovative redesign of both the numerical model and the information infrastructure where it is nested within.
The traditional working method for modelers is to prepare the input of the schematization (grid and input measurements), run a model, wait a while for the model to complete and finally post-process the results. This is working method is suitable for confirmatory modeling (making the best prediction a few days ahead). Most of the questions that people use numerical models for are exploratory. To effectively plan and design a water system it is important to provide an indication of the effect of a measure in seconds, not over night or even over coffee.
The traditional working method was also that the modeler was in control. The exploratory approach is often a social process where multiple stakeholders interact and the model is used to test and evaluate ideas. This requires the stakeholders to be in control and not the modeler. The main visualization and interaction method is through a mobile compatible website.
The amount of data involved in preparing the model is rapidly increasing. There is an increase in supply of data caused by the ever growing measurement resolution and spatial coverage. For example the topology of waterbodies at a spatial coverage of a small country are used to determine if the water will find its way into ones neighborhood. The topography and infiltration rate at a spatial resolution of a planter determine if the water will flow to ones backyard or the neighbors.
This resulted in the question how can we design a platform that is suitable for this new modeling approach and meets the corresponding challenges.
The 3Di platform, in which the new approach was applied, consists of a new overland flooding model (1D/2D) and a corresponding infrastructure that has been applied for both exploratory and confirmatory modeling of rural and urban areas at coverages of O(1000km2) at a resolution of O(0.1m).
The advances on a numerical level, by implementing a quad tree structure allowing only the relevant physics to be computed, resulting in an order of magnitude speedup have been previously presented. Here we present the new architecture and it's underlying components, resulting in the advances in model interaction and scalability.
The new multi-tier infrastructure was designed under the following conditions:
Every aspect of the system should be 100 times faster than the previous approach.
The main end user is the water-manager, who is assumed to only know how to use a browser and likes to present realistic looking results.
The modeling should be interactive and social, meaning that multiple people can interact with the same running model at the same time.
The system should be able to run both in private and public clouds. The numerical model should be able to run on a single computer on a supercomputer and in the cloud.
The main interface is a mobile compatible website. The model results are rendered on the client in a html5 web map using both bitmap and vector graphics. The user can interact with the model in several ways, for example by changing properties of hydraulic structures, water levels, discharge locations, rain events. The browser transforms these interactions into messages which are pushed back to the model through websockets. Model results are also pushed to the browser using websockets and on request through a Web Map Service (WMS) extended with time support.
The web application server forwards the websockets messages back and forth between the browser and the model. This application also keeps track of authorization, authentication, user preferences and the social aspects, who is watching which model, who is in control.
The numerical model was adapted to an interactive model by implementing the low level, non invasive BMI interface. This interface makes the numerical model adhere to the Hollywood principle (don't call us, we'll call you): the control of the initialize, update and finalize steps are called from outside of the numerical model. All variables of the model can be accessed through direct or shared memory access, allowing for communication within nanoseconds. The numerical model can be run as a command line executable (for operational runs), in a GUI (for schematization design), as a shared library (for data exchange) and as a message publisher/puller (for interaction). A separate tool generates photo-realistic renderings in 3D stereo for presentation purposes. The messaging publisher module, which works for any BMI compliant model, allows to push results to the browser at a rate of over 20 time per second and a latency of under 50ms.
A provisioning module is used to quickly deploy new updates of models schematization and to scale when more running models or WMS servers are required. Models can run at cloud providers and at the central server. Model schematizations are kept under distributed version control so that updates can be pushed and different versions of the same model can be used.
The new platform has been deployed at several locations in the Netherlands and in Singapore. In preliminary tests we found that the model has skill comparable to previous models. Setting the bar for performance a few order of magnitudes higher resulted in new ideas instead of improved existing ideas. A major part of the platform (model messaging, WMS server) has been released open source because it can also be used with other models. The numerical model is shared within a small community.
By solving the challenges new challenges arise. The new challenges are technical and social. How can we dynamically filter out the most relevant information from the platform as the bandwidth and browsers rendering power of the mobile browsers is limited and varying. The shift towards social and exploratory modeling and away from the modeler "in control" require new best practices. How do we make sure information is used correctly? How "true" is the information perceived?

\section{Method}

Header of the first page should be as appears in this instructions' paper (right justified). Headers of following pages should be: the name of the authors (initial(s) and name) / title of the paper (centered). If there are more than two authors write first author et al. If the title is too long, cut the title where needed and put three dots. Headers text should be in italics 8 pt. Arial.
No footer should appear. \textbf{Do not} number the pages. Numbering of the pages will be done by the editors of the Proceedings.

\section{Results}
Equations should be numbered consecutively as they appear in the
text with Arabic numerals and should be referred to by their
numbers only, e.g. (3).  Equations must be typed not hand printed.
About 5 mm should be left clear above and below each equation.

\begin{figure}[h]
\centering
\includegraphics[width=10cm]{interpolationlevees2}
\caption{Interpolation method}
\label{fig1}
\end{figure}

\section{Conclusions and Recommendations}

It was shown by \citet{berry:teomowcihbsi} that...

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\bibliographystyle{iemss}
\bibliography{iemss}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\end{document}

### 2014-03-07 article.tex
writing

@@
 \usepackage[pdftex]{graphicx}
-\usepackage{natbib}
-\usepackage{amsmath}
+\usepackage[authoryear]{natbib}
 \usepackage[labelfont=bf,textfont=sf,labelsep=period]{caption}
 \usepackage[english]{babel}
-%\RequirePackage{lineno} % for line numbers
+% \RequirePackage{lineno} % for line numbers
 
@@
 
-%Authors Names and Affiliations:  Two spaces below the title, 10 pt
-%Arial, Upper and Lower Case, underline author presenting
-%paper and  provide his/her email address
+% Authors Names and Affiliations:  Two spaces below the title, 10 pt
+% Arial, Upper and Lower Case, underline author presenting
+% paper and  provide his/her email address
 
@@
 Several challenges required an innovative redesign of both the numerical model and the information infrastructure where it is nested within.
+
+These challenges arise from the movement towards more efficient working methods.
+\begin{itemize}
+\item \emph{Confirmatory versus exploratory}  \cite{Baart2013}
+\item \emph{Passive versus interactive}
+\item \emph{Solitary versus social}
+\end{itemize}
+
+Other challenges arise from the increased data volumes.
+\begin{itemize}
+  \item \emph{Measurements}
+  \item \emph{Communication}
+\end{itemize}
+
 The traditional working method for modelers is to prepare the input of the schematization (grid and input measurements), run a model, wait a while for the model to complete and finally post-process the results.
@@
 Most of the questions that people use numerical models for are exploratory.
-To effectively plan and design a water system it is important to provide an indication of the effect of a measure in seconds, not over night or even over coffee.
+
 The traditional working method was also that the modeler was in control.
@@
 The main visualization and interaction method is through a mobile compatible website.
+
 The amount of data involved in preparing the model is rapidly increasing.
@@
 This resulted in the question how can we design a platform that is suitable for this new modeling approach and meets the corresponding challenges.
+
 The 3Di platform, in which the new approach was applied, consists of a new overland flooding model (1D/2D) and a corresponding infrastructure that has been applied for both exploratory and confirmatory modeling of rural and urban areas at coverages of O(1000km2) at a resolution of O(0.1m).
@@
 \section{Method}
-
-Header of the first page should be as appears in this instructions' paper (right justified).
-Headers of following pages should be: the name of the authors (initial(s) and name) / title of the paper (centered).
-If there are more than two authors write first author et al.
-If the title is too long, cut the title where needed and put three dots.
-Headers text should be in italics 8 pt.
-Arial.
-No footer should appear. \textbf{Do not} number the pages. Numbering of the pages will be done by the editors of the Proceedings.
+\subsection{}
+\subsection{}
 
 \section{Results}
-Equations should be numbered consecutively as they appear in the
-text with Arabic numerals and should be referred to by their
-numbers only, e.g. (3).  Equations must be typed not hand printed.
-About 5 mm should be left clear above and below each equation.
-
+\subsection{Numerical model}
+\subsection{Architecture}
+\subsection{Communication}
+\subsection{Visualization}
 \begin{figure}[h]
@@
 
-It was shown by \citet{berry:teomowcihbsi} that...
-
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 \bibliographystyle{iemss}
-\bibliography{iemss}
+\bibliography{bibliography}
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

### 2014-03-14 article.tex
article
reorganize
writing stuff

@@
 \usepackage[english]{babel}
+
+% extra packages not in the template
+% Don't color hyperlinks
+\usepackage{wrapfig}
+\usepackage{paralist}
+\usepackage[draft]{hyperref}
+\usepackage{amsmath}
+\usepackage[printonlyused,withpage]{acronym}
+\usepackage{sansmath}
+
 % \RequirePackage{lineno} % for line numbers
@@
 % Text to appear in the header of the pages
-\IEMSShead{Baart et al. / Interactive web based flood modeling}
+\IEMSShead{Baart et al. / Interactive web-based flood modeling}
 
@@
       Delft, The Netherlands (fedor.baart@deltares.nl, arthur.vandam@deltares.nl, gennadii.donchyts@deltares.nl)}},
-  J. Ha
+  K. K. Ha
   \address[B1]{\it{Nelen \& Schuurmans,
@@
 \begin{document}
+% make sure everything is sans-serif, including math
+\sffamily
+\sansmath
 
@@
 \section{Introduction}
-
 The flooding of rural and urban areas is an increasing hazard to society.
@@
 
-These challenges arise from the movement towards more efficient working methods.
+These challenges arise from the movement towards more modern working methods and the ever increasing data volumes. The working methods are changing on different aspects.
 \begin{itemize}
-\item \emph{Confirmatory versus exploratory}  \cite{Baart2013}
-\item \emph{Passive versus interactive}
-\item \emph{Solitary versus social}
+\item \emph{Mono-disciplinary versus integrated} Models used to be created by one or few people working in the same discipline.
+The challenge now is to model the integrated environment, which can best be approached as a community \citep{Voinov2010} of different disciplines.
+\item \emph{Confirmatory versus exploratory} From assuming that the model is valid and can be used for hypothesis testing to realizing it's a made-up simplification of reality \citep{Oreskes1994} that can be used to explore possibilities.
+\item \emph{Passive versus interactive} Moving from the pre-processing, wait a few hours, post-processing cycle towards a millisecond feedback loop with high resolution and connected physical processes.
+The challenge here is to move from the scale of a square kilometer \citep[for example][]{Losasso2008} to the scale of small country.
+\item \emph{Solitary versus social} The solitary modeler creating static pictures for a 600 page report describing all the different scenarios to a group of stakeholders in control \footnote{preferably with guidance from domain experts} of a model running simultaneous on their mobile devices and a big screen in the control room.
 \end{itemize}
-
-Other challenges arise from the increased data volumes.
-\begin{itemize}
-  \item \emph{Measurements}
-  \item \emph{Communication}
-\end{itemize}
-
-The traditional working method for modelers is to prepare the input of the schematization (grid and input measurements), run a model, wait a while for the model to complete and finally post-process the results.
-This is working method is suitable for confirmatory modeling (making the best prediction a few days ahead).
-Most of the questions that people use numerical models for are exploratory.
-
-The traditional working method was also that the modeler was in control.
-The exploratory approach is often a social process where multiple stakeholders interact and the model is used to test and evaluate ideas.
-This requires the stakeholders to be in control and not the modeler.
-The main visualization and interaction method is through a mobile compatible website.
 
@@
 
-The 3Di platform, in which the new approach was applied, consists of a new overland flooding model (1D/2D) and a corresponding infrastructure that has been applied for both exploratory and confirmatory modeling of rural and urban areas at coverages of O(1000km2) at a resolution of O(0.1m).
-The advances on a numerical level, by implementing a quad tree structure allowing only the relevant physics to be computed, resulting in an order of magnitude speedup have been previously presented.
+% Several attempts have been made to enable the movement towards these new working methods and increased data volumes.
+% Refs Numerical (adaptive mesh): overview
+% Interactive models: IRF: ESMF, OpenMI
+% Social modelling:
+% Web based modelling:
+
+The \threedi  platform, in which these approaches and several new ones are applied, consists of a new overland flooding model (1D/2D) and a corresponding infrastructure that has been applied for interactive web-based modeling of rural and urban areas at coverages of \orderof(1000km2) at a resolution of \orderof(0.1m).
+Here we describe the advances in the numerical methods, architecture and visualization techniques that were used to create the platform.
+
+% The advances on a numerical level, by implementing a quad tree structure allowing only the relevant physics to be computed, resulting in an order of magnitude speedup have been previously presented.
 Here we present the new architecture and it's underlying components, resulting in the advances in model interaction and scalability.
-The new multi-tier infrastructure was designed under the following conditions:
-Every aspect of the system should be 100 times faster than the previous approach.
-The main end user is the water-manager, who is assumed to only know how to use a browser and likes to present realistic looking results.
-The modeling should be interactive and social, meaning that multiple people can interact with the same running model at the same time.
-The system should be able to run both in private and public clouds. The numerical model should be able to run on a single computer on a supercomputer and in the cloud.
-The main interface is a mobile compatible website.
+% The new multi-tier infrastructure was designed under the following conditions:
+% Every aspect of the system should be 100 times faster than the previous approach.
+% The main end user is the water-manager, who is assumed to only know how to use a browser and likes to present realistic looking results.
+% The modeling should be interactive and social, meaning that multiple people can interact with the same running model at the same time.
+% The system should be able to run both in private and public clouds. The numerical model should be able to run on a single computer on a supercomputer and in the cloud.
+% The main interface is a mobile compatible website.
 The model results are rendered on the client in a html5 web map using both bitmap and vector graphics.
@@
 Model results are also pushed to the browser using websockets and on request through a Web Map Service (WMS) extended with time support.
-The web application server forwards the websockets messages back and forth between the browser and the model.
+% The web application server forwards the websockets messages back and forth between the browser and the model.
 This application also keeps track of authorization, authentication, user preferences and the social aspects, who is watching which model, who is in control.
-The numerical model was adapted to an interactive model by implementing the low level, non invasive BMI interface.
+% The numerical model was adapted to an interactive model by implementing the low level, non invasive BMI interface.
 This interface makes the numerical model adhere to the Hollywood principle (don't call us, we'll call you): the control of the initialize, update and finalize steps are called from outside of the numerical model.
@@
 The messaging publisher module, which works for any BMI compliant model, allows to push results to the browser at a rate of over 20 time per second and a latency of under 50ms.
-A provisioning module is used to quickly deploy new updates of models schematization and to scale when more running models or WMS servers are required.
+% A provisioning module is used to quickly deploy new updates of models schematization and to scale when more running models or WMS servers are required.
 Models can run at cloud providers and at the central server.
 Model schematizations are kept under distributed version control so that updates can be pushed and different versions of the same model can be used.
-The new platform has been deployed at several locations in the Netherlands and in Singapore.
+% The new platform has been deployed at several locations in the Netherlands and in Singapore.
 In preliminary tests we found that the model has skill comparable to previous models.
@@
 The numerical model is shared within a small community.
-By solving the challenges new challenges arise.
+% By solving the challenges new challenges arise.
 The new challenges are technical and social.
@@
 
-\section{Method}
-\subsection{}
-\subsection{}
+\section{Numerical model}
+The computational core originates from core components of the {SOBEK} 1D2D flooding software \citep{Stelling2006}, and has undergone three major transformations. \begin{inparaenum}
+\item a streamlining of the software architecture to offer access to the memory of a running model,
+\item overland flow at subgrid scale and
+\item inclusion of hydrological processes in the subsoil also at subgrid scale.
+\end{inparaenum}
 
-\section{Results}
-\subsection{Numerical model}
-\subsection{Architecture}
-\subsection{Communication}
-\subsection{Visualization}
+The traditional standalone flooding program and its integrated \ac{GUI} have been split into dedicated components.
+Most importantly, the core library with all numerical algorithms and I/O.
+On top of this core library are independent ways of addressing the core: a light-weight interface library that exposes the \ac{API}, and the original technical \ac{GUI} that acts as a sandbox for early developments.
+The core and \ac{API} offer powerful new functionality for direct and fast interaction using introspection (described in \autoref{sec:coupling} and the Hollywood-principle.
+
+By introspection, the outside world (e.g., a Python-based website, a C\#-\ac{GUI}, a MATLAB-script) can examine model state at run time.
+More specifically: get current water levels and flow fields, current running state of water pumps and other structures, active meteorological forcings.
+Each of these state variables can just as easily be \emph{set} as they are \emph{read}.
+The \ac{API} exposes memory pointers to relevant model state variables.
+The model will directly 'feel' and use its new state.
+
+The freedom of changing model state is possible due to the core and \ac{API} adhering to the Hollywood Principle: ``Don't call us, we'll call you''.
+The core or model does not run itself, but instead is instructed by the outside world.
+Typically, the timeloop from a conventional program is extracted and moved to the calling application; only the single timestep remains in the core.
+
+In several workshops, model coupling frameworks \citep[see][for an overview]{Jagers2010}, have agreed on the minimal interface that makes a model adhere to the Hollywood principles, by exposing the following methods:
+
+\begin{enumerate}
+\item \textbf{initialize()}: Load and initialize a model from file.
+\item \textbf{run($\Delta t$)}: Run a single timestep.
+\item \textbf{finalize()}: Finalize the model, memory cleanup, file flushing, statistics reporting.
+\end{enumerate}
+
+In between these interface calls, the model state may be changed from outside and the core should support that.
+The library is loaded and can remain a running process for a long time, for many different models..
+A positive side-effect of implementing this interface is that it may reveal subtle inconsistencies in existing code, for example when initializing different models after one another: incomplete cleanup of previous settings, already allocated memory, unclosed file pointer.
+
+The power and elegance of the subgrid concept is now also proving itself for the hydrological processes in the subsoil.
+The planter size pixel resolution of the digital terrain maps for detailed evaluation of overland flow friction and water storage is described in \citet{Stelling2012}.
+Coupling the 2D grid with 1D channels and controllable structures enables greater detail in civilized regions.
+
+Another important influence on potential flooding is the storage of water in the subsoil, or the lack thereof. The change in surface water volume is driven by several familiar processes:
+\[
+\Delta V_\mathrm{surface\ water}=(Q_\mathrm{overland} + Q_\mathrm{precipitation} - Q_\mathrm{infiltration} - Q_\mathrm{evapo(transpi)ration} + Q_\mathrm{drainage})\cdot\Delta t
+\]
+Infiltration, evaporation and drainage are now also evaluated at pixel resolution, whereas the computational grid cells are larger and ensure high performance.
+
+The heterogeneous landscape and land use (be it it urban and/or rural) are specified in the model as gigapixel bitmaps for soil types, crop types, drainage resistance and maximum infiltration.
+Consider the basic head-dependent linear infiltration in (\ref{eq:qinf}) as an example.
+Note that this specific type of head-dependent infiltration rate in can be chosen freely, but the fact that it uses pixel-resolution soil characteristics is what is important here.
+%
+\begin{figure}[h]
+\begin{minipage}[b]{0.55\linewidth}
+\begin{multline}
+\label{eq:qinf}
+Q_\mathrm{infiltration}=\\
+\sum_{\mathrm{pixels}\ i,j} \frac{\min(0, \zeta_\mathrm{sw}-\max(h_\mathrm{gw}, d_{i,j}))}{\mathrm{resistance}_{i,j}}\Delta x_p \Delta y_p,
+\end{multline}
+where $\zeta_\mathrm{sw}$ and $h_\mathrm{gw}$ are surface and ground water level, respectively, and $d_{i,j}$ is the surface/bed level of pixel $(i,j)$.
+Notice how dry pixels (higher land) do not amount to the total infiltration.
+
+\end{minipage}
+\hspace{0.5cm}
+\begin{minipage}[t]{0.4\linewidth}
+\centering
+\includegraphics[width=\textwidth]{subgrid_hydrology}
+\caption{Subgrid infiltration into heterogeneous landscape.}
+\label{fig:figure2}
+\end{minipage}
+\end{figure}
+%
+These advances in the numerical model allow the model to act as a component in the interactive web-based model environment that will be described in the next section.
+
+\section{Architecture}
+\begin{wrapfigure}[12]{r}{0.5\textwidth}
+  \centering
+  \includegraphics[width=0.48\textwidth]{webgui}
+  \caption{Web \ac{GUI}}
+  \label{fig:gui}
+\end{wrapfigure}
+
+To enable social and interactive modeling through a map environment the most logical choice was to use a web based environment.
+Over the last years this choice has become easier through the large list of advancements for browser based applications.
+
+The end-users, interact with the web based \ac{GUI}.
+The design philosophy is minimalistic, allowing little room for user mistakes.
+For example manholes can only be placed inside the model domain, and can be controlled by only two different discharge settings (normal and extreme discharge).
+It was tested in several sessions with end-users that this interface was the most usable interface for the intended audience.
+
+\subsection{Overview}
 \begin{figure}[h]
   \centering
-  \includegraphics[width=10cm]{interpolationlevees2}
-  \caption{Interpolation method}
-  \label{fig1}
+  \includegraphics[width=10cm]{architecture}
+  \caption{Architecture overview}
+  \label{fig:arch}
 \end{figure}
 
-\section{Conclusions and Recommendations}
+\subsection{Web based models }
+\begin{itemize}
+  \item provisioning
+  \item interaction
+  \item social modeling
+\end{itemize}
+\subsection{Data sources }
+\begin{itemize}
+  \item lidar
+  \item rain example
+  \item structures/levees
+\end{itemize}
+
+\section{Coupling}
+\label{sec:coupling}
+
+\begin{itemize}
+  \item raster
+  \item quad unstructured
+  \item 1d network
+  \item structure tables
+\end{itemize}
+
+\subsection{BMI}
+
+Basic Model Interface Wrapper
+The \ac{BMI} is a low level API designed access both control and data access functions of the model engine component.
+It is similar to the type introspection used in many programming languages, however, instead of the class types \ac{BMI} provides access to the model variables, including their attributes such as data type, unit of measure, input or output role.
+To provide full access to the internals of the internal model state the following function types should be implemented: model control, model information, variable information, variable getter and setter and grid information.
+The \ac{BMI} API is quite compact and does not require much effort to implement it in the model.
+However, it can be quite time consuming to continuously extend access to the model variables in programming languages such as fortran.
+To simplify this maintenance process we have implemented an automated way to add or remove model variables via \ac{BMI}.
+The automation was added using Python script and Mako template engine (Figure 1).
+
+\begin{figure}[h]
+  \centering
+  \includegraphics[width=10cm]{bmi_3di}
+  \caption{Basic Model Interface template}
+  \label{fig:bmi}
+\end{figure}
+
+Figure 1: Automatic generation of the include files for \ac{BMI} support
+
+The Mako template files used to generate FORTRAN include all code necessary to expose all required external functions of the model engine shared library.
+This library is then used by the 3Di model wrapper (developed using Python and ctypes) to control model.
+
+The approach described above significantly decreased costs of change required to maintain \ac{BMI} support in the 3Di model.
+
+Additionally to the standard functions we had to extend BMI with support for simple compound data types (C structures using only simple fields).
+This was necessary to simplify access to the higher level model entities such as hydraulic structures (e.g.
+pumps, weirs, culverts).
+
+\subsection{MMI}
+\begin{itemize}
+  \item message interface
+  \item serialization
+  \item comparison to other interfaces
+\end{itemize}
+\subsection{Websockets}
+\begin{itemize}
+  \item Advantages of push for models
+  \item Serialization
+\end{itemize}
+\section{Visualization}
+\subsection{Ravensburger }
+\begin{itemize}
+  \item performance comparison
+  \item WMS/TWMS coupling
+\end{itemize}
+\subsection{Interpolation}
+Interpolation Martijn
+
+The 3Di platform enables a more exploratory approach to modeling, serving a wide variety of end-users.
+The variety of end-users exposes the numerical model to non-technical users with little understanding of implemented numerical schemes.
+Some artefacts resulting from the numerical schemes are therefore filtered and smoothed using an interpolation method.
+Artefacts include squared water depth visualization and water present on two sides of a levee within one calculation cell.
+The filtering and smoothing reduces the need for prior knowledge of numerical modeling.
+
+The numerical core of the 3Di platform is based on a subgrid calculation method (Stelling, 2012).
+The hydrodynamic calculations are performed on a non-uniform grid (quadtree ordering), with use of high resolution \ac{DEM} information as input.
+Flow variables, such as water levels and flow velocities, are available on the calculation grid, which reduces storage costs.
+The resulting adaptive storage of flow variables results in artefacts after plotting them on the high resolution \ac{DEM} rasters.
+Figure 1a shows the occurring artefacts.
+
+\begin{figure}[h]
+  \centering
+  \includegraphics[width=0.5\textwidth]{artifacts_a}
+  \caption{Interpolation method}
+  \label{fig:artifacts_a}
+\end{figure}
+
+\begin{figure}[h]
+  \centering
+  \includegraphics[width=0.5\textwidth]{artifacts_b}
+  \caption{Interpolation method}
+  \label{fig:artifacts_b}
+\end{figure}
+
+Figure 1a.  Artefacts in the visualization                      Figure 1b.  Filtered visualization
+
+A barycentric linear interpolation technique is selected for filtering artefacts on the subgrid level, because of two main restrictions.
+First, the web visualization requires a 50 ms rendering time in the browser that is still obtained using this method.
+Second, the ability to interpolate water level results on an arbitrary grid.
+The quadtree calculation grid was triangulated from the cell centers using existing software libraries, creating a triangulated mesh (Figure 2b).
+Water level information is present on cell centers and is then linearly interpolated to the underlying subgrid.
+The interpolation uses a weighted average the determine water levels on the subgrid.
+The determination of the triangular mesh is very low cost after which the interpolation can be executed immediately.
+The little calculation steps needed result in high performance.
+This technique has a large upside in calculation time compared to a technique using cell corners as interpolation points.
+When using cell corners, an extra intermediate calculation has to be made to determine water levels at cell corners.
+After the extra calculation the interpolation is executed.
+Results of plotting water depths after the barycentric linear interpolation are shown in Figure 1b.
+
+The triangulation within the linear barycentric interpolation also allows us to filter levee artefacts of the numerical model .The calculation grid incorporates the function of a levee in the numerical schemes.
+However it disregards the actual position of such a levee within a calculation cell.
+The triangulation can reconstruct levees with use of additional levee geometries.
+A triangle is one of the simplest geometrical forms and can be used reconstruct any arbitrary shape (Welch and Witkin, 1994).
+To incorporate the levee in the trianglulation, the levee and grid geometries are intersected (Figure 2a).
+For the intersected calculation cells the influential water level points are determined and coupled to the relevant calculation areas.
+The arrows in figure 2a indicate which water level is relevant for the dike intersected cell areas.
+When all waterlevel points are allocated to the dike, the triangulation can be executed with the incorporation of the levee geometries.
+The newly created triangular mesh now also follows the added levee geometry.
+Successively the interpolation is executed, resulting in water being retained behind the dike within one calculation cell.
+The interpolated result is shown in Figure 2b.
+
+\begin{figure}[h]
+  \centering
+  \includegraphics[width=0.5\textwidth]{interpolationlevees3}
+  \caption{Interpolation method}
+  \label{fig:interp_a}
+\end{figure}
+
+Figure 2a. Levee and grid geometry intersections        Figure 2b. Levee incorporated interpolation
+
+Reference:
+Welch, W.
+Witkin, A.
+Free-Form Shape Design Using Triangulated Surfaces, Proceeding SIGGRAPH '94 Proceedings of the 21st annual conference on Computer graphics and interactive techniques, Pages 247-256, 1994
+Stelling, G.S. Quadtree flood simulations with sub-grid digital elevation models, Proceedings of the ICE - Water Management, Volume 165, Issue 10, 01 November 2012, pages 567 –580
+
+http://quietplease.com/home/tri.pdf
+
+\section{Limitations: \emph{All}}
+Several other models use similar approaches, albeit with limitations.
+The subgrid scheme by \citet{Stelling2012} is partly based on the subgrid method for flooding and drying by \citet{Casulli2009}, who implements his methods in the UnTRIM research package \citet{Casulli2000}.
+UnTRIM uses fully unstructured grids as opposed to the quad tree grids used here.
+The coupled 1D-2D hydrodynamic model MIKE-FLOOD \citep{Dhi2014} also offers subgrid overland flow.
+In the accompanying MIKE-SHE all relevant vertical hydrological processes are offered, also with some degree of subgrid-information for the infiltration part.
+
+\section{Future plans: \emph{All}}
+In that respect, combining the subgrid approach with our own unstructured flow solver Delft3D-FLOW Flexible Mesh \citep{Kernkamp2011} which is already equipped with an interactive BMI-interface) is a logical future step.
+\section{Conclusions: \emph{Fedor}}
 
@@
 
+\begin{acronym}[AAAAA]
+\acro{H2O}[$\mathrm{H_2O}$]{water}
+\acro{BMI}{Basic Model Interface}
+\acro{API}{Application Programming Interface}
+\acro{GUI}{Graphical User Interface}
+\acro{DEM}{Digital Elevation Model}
+\end{acronym}
+
 \end{document}

### 2014-03-20 article.tex
writing

@@
 
+\begin{wrapfigure}[19]{r}{0.5\textwidth}
+\centering
+\includegraphics[width=0.48\textwidth]{subgrid_hydrology}
+\caption{Subgrid infiltration into heterogeneous landscape.}
+\label{fig:figure2}
+\end{wrapfigure}
+
 In between these interface calls, the model state may be changed from outside and the core should support that.
@@
 Consider the basic head-dependent linear infiltration in (\ref{eq:qinf}) as an example.
-Note that this specific type of head-dependent infiltration rate in can be chosen freely, but the fact that it uses pixel-resolution soil characteristics is what is important here.
 %
-\begin{figure}[h]
-\begin{minipage}[b]{0.55\linewidth}
-\begin{multline}
+\begin{equation}
 \label{eq:qinf}
-Q_\mathrm{infiltration}=\\
-\sum_{\mathrm{pixels}\ i,j} \frac{\min(0, \zeta_\mathrm{sw}-\max(h_\mathrm{gw}, d_{i,j}))}{\mathrm{resistance}_{i,j}}\Delta x_p \Delta y_p,
-\end{multline}
+Q_\mathrm{infiltration}= \sum_{\mathrm{pixels}\ i,j} \frac{\min(0, \zeta_\mathrm{sw}-\max(h_\mathrm{gw}, d_{i,j}))}{\mathrm{resistance}_{i,j}}\Delta x_p \Delta y_p,
+\end{equation}
 where $\zeta_\mathrm{sw}$ and $h_\mathrm{gw}$ are surface and ground water level, respectively, and $d_{i,j}$ is the surface/bed level of pixel $(i,j)$.
@@
 
-\end{minipage}
-\hspace{0.5cm}
-\begin{minipage}[t]{0.4\linewidth}
-\centering
-\includegraphics[width=\textwidth]{subgrid_hydrology}
-\caption{Subgrid infiltration into heterogeneous landscape.}
-\label{fig:figure2}
-\end{minipage}
-\end{figure}
+Note that this specific type of head-dependent infiltration rate in can be chosen freely, but the fact that it uses pixel-resolution soil characteristics is what is important here.
+
 %
@@
 \section{Architecture}
-\begin{wrapfigure}[12]{r}{0.5\textwidth}
+\begin{wrapfigure}[12]{l}{0.5\textwidth}
+  \centering
+  \includegraphics[width=0.48\textwidth]{architecture}
+  \caption{Architecture overview}
+  \label{fig:arch}
+\end{wrapfigure}
+
+To enable social and interactive modeling through a map environment the most logical choice was to use a web based environment.
+Over the last years this choice has become easier through the large list of advancements for browser based applications.
+An overview of the architecture is given in \autoref{fig:arch}.
+Notice how the architectural follows the general Model View Controller pattern \citep{GOF2004}.
+On a high level the architecture follows is classic.
+In this section we will describe the lower level, yet important,  technical innovations that made the application of this architecture successful.
+
+\begin{wrapfigure}[16]{r}{0.5\textwidth}
   \centering
@@
 
-To enable social and interactive modeling through a map environment the most logical choice was to use a web based environment.
-Over the last years this choice has become easier through the large list of advancements for browser based applications.
-
 The end-users, interact with the web based \ac{GUI}.
@@
 For example manholes can only be placed inside the model domain, and can be controlled by only two different discharge settings (normal and extreme discharge).
-It was tested in several sessions with end-users that this interface was the most usable interface for the intended audience.
+Sessions with groups of end-users resulted in the idea that this interface was the most usable.
+The design goals of providing freedom to explore together with high resolution high frequency model output, yet without the insinuation that the model results are ``true'' proves to be challenging.
+Many end-users confuse high resolution, detailed tuning parameters and higher frequency results with higher accuracy.
 
-\subsection{Overview}
-\begin{figure}[h]
-  \centering
-  \includegraphics[width=10cm]{architecture}
-  \caption{Architecture overview}
-  \label{fig:arch}
-\end{figure}
+Several challenges arise when preparing a model for a web-based use.
+One of the differences when comparing a desktop user to a web-based user is the short attention span.
+If no indication of progress is given for longer than two second users will start reloading, clicking erratically or will just close the website as a whole searching for a new ``frisson'' \citep{Carr2011}.
 
-\subsection{Web based models }
-\begin{itemize}
-  \item provisioning
-  \item interaction
-  \item social modeling
-\end{itemize}
+To make sure that the model is ready to run when a user arrives and no time is lost with starting up, initializing, reading files a lot of attention is given to the provisioning of the model.
+Provisioning in this context is preparing a model controller so that it starts it's computation on request.
+Machines are prepared and put on hold at one of the \ac{IaaS} providers (such as Amazon).
+These machines are prepared with a container that contains the model schematization, tables and grids and hot restart files.
+The tables, grids and restart files are read into memory so that the model is ready to start its computation when it is needed.
+This results in a time between a user logging in and a model start below 5 seconds.
+
+Interaction is another challenge where timing is crucial.
+When a user interacts with the model, for example places a discharge point, turns on a pump or lowers an orifice she expects a response within a second.
+This short feedback loop is achieved using several techniques.
+Using the regular \ac{HTTP} GET Request is too slow in this case.
+To reduce communication we use a chain of direct connections.
+This starts with a websocket \citep{w3c2010} connection between the message broker and the browser.
+The model controller gets notified directly of user interactions.
+These interactions are collected in a separate thread and pushed into the model on every timestep.
+These timesteps have to be kept shorter than in a classical model run.
+A timestep resulting in a computation time of \si{100}{\mili\second} is common.
+After a model timestep the chain is inverted.
+The model pushes it's arrays back in the direction of the browser, which is subscribed to these events and can directly visualize the results.
+This message broker approach allows interactions with a feedback time in the order of \si{100}{\mili\second}, which gives a good sense of interactivity.
+
+As a final challenge there is the social aspect of modeling.
+Several group sessions with different settings have been tried.
+The current setup is that the model is run in a channel, like in a chat room.
+Multiple people can join the model run.
+One person is in control and others can view and zoom in on the same model run.
+Other users can request control of the model.
+The sessions worked best when a topic expert, not necessary a modeler was present in the session.
+
 \subsection{Data sources }
@@
 \acro{DEM}{Digital Elevation Model}
+\acro{HTTP}{Hyper Text Transfer Protocol}
+\acro{IaaS}{Infrastructure as a Service}
 \end{acronym}

### 2014-03-21 figures/arch.tex
figure

\documentclass{article}
\usepackage[latin1]{inputenc}
\usepackage{tikz}
\usetikzlibrary{calc,shapes,arrows,shadows,positioning,backgrounds,fit} %,external
%\tikzexternalize
\usepackage[active,pdftex,tightpage]{preview}
\PreviewEnvironment{tikzpicture}
\begin{document}
\sffamily
\pagestyle{empty}
\renewcommand{\rmdefault}{phv} % Arial
\renewcommand{\sfdefault}{phv} % Arial

\tikzstyle{component} = [diamond, draw,text badly centered,inner sep=0pt]
\tikzstyle{block} = [rectangle, draw,text centered, rounded corners,minimum size=1cm]
\tikzstyle{line} = [draw, -latex']
\tikzstyle{cloud} = [draw, ellipse]
\tikzstyle{nodeEdge} = [black,thick,->];
\tikzstyle{nodeBi} = [black,thick,<->];

\def\blockdist{2.3}

\begin{tikzpicture}[text width=6em,auto]

    % Place nodes
    \node [block] (core) {numerical engine};
    \node [block] (controller) [right=0.5cm of core] {model controller};
    \node [block] (schema) [below=0.5cm of core] {schematization};
    \node [block] (feed) [above=1cm of controller] {data feed};
    \node [block] (broker) [below right=0cm and 0.5cm of controller] {event+action\\ cache};
    \node [block] (cache) [above right=0cm and 0.5cm of controller] {data cache};
    \node [block] (maps) [right=0.5cm of cache] {time+map \\ server};
    \node [block] (sockets) [right=0.5cm of broker] {websocket server};
    \node [block] (browser) [below right=0 cm and 1cm of maps] {browser};

    % Connections
    \draw ($(schema.north)$) edge[nodeEdge] ($(core.south) $);
    \draw ($(core.east) $) edge[nodeBi] ($(controller.west)$);
    \draw ($(feed.south) $) edge[nodeEdge] ($(controller.north)$);
    \draw ($(feed.south) $) edge[nodeEdge] ($(controller.north)$);
    \draw ($(controller.north) $) edge[nodeEdge,out=90,in=180] ($(cache.west)$);
    \draw ($(controller.south) $) edge[nodeBi,out=270,in=180] ($(broker.west)$);
    \draw ($(cache.east) $) edge[nodeEdge] ($(maps.west)$);
    \draw ($(broker.east) $) edge[nodeBi] ($(sockets.west)$);
    \draw ($(maps.east) $) edge[nodeEdge,out=0,in=90] ($(browser.north)$);
    \draw ($(sockets.east) $) edge[nodeBi,out=0,in=270] ($(browser.south)$);

\end{tikzpicture}
\end{document}

### 2014-03-21 article.tex
writing

@@
 
-%
 These advances in the numerical model allow the model to act as a component in the interactive web-based model environment that will be described in the next section.
@@
 To enable social and interactive modeling through a map environment the most logical choice was to use a web based environment.
-Over the last years this choice has become easier through the large list of advancements for browser based applications.
+Over the last years this choice has become easier.
+The large list of browser enhancements (for example inline \ac{SVG}, canvas, plugin free video \citep{Berjon2014} and \ac{CSS} animations \citep{Jackson2013}) allow for rich visualizations.
 An overview of the architecture is given in \autoref{fig:arch}.
-Notice how the architectural follows the general Model View Controller pattern \citep{GOF2004}.
+Notice how the architectural follows the general Model View Controller pattern \citep{Gamma1994}.
 On a high level the architecture follows is classic.
@@
 \begin{acronym}[AAAAA]
+\acro{CSS}{Cascading Style Sheets}
+\acro{SVG}{Scalable Vector Graphics}
 \acro{H2O}[$\mathrm{H_2O}$]{water}

### 2014-03-21 article.tex
spelling

--- 2014-03-20 article.tex
@@
 % Don't color hyperlinks
+\usepackage{siunitx}
 \usepackage{wrapfig}
@@
 \usepackage{amsmath}
-\usepackage[printonlyused,withpage]{acronym}
+\usepackage[printonlyused]{acronym}
 \usepackage{sansmath}
+\usepackage{listings}
+\usepackage[list=true]{subcaption}
 
 % \RequirePackage{lineno} % for line numbers
+
+\usepackage{color}
+
+\definecolor{mygreen}{rgb}{0,0.6,0}
+\definecolor{mygray}{rgb}{0.2,0.2,0.2}
+\definecolor{mymauve}{rgb}{0.58,0,0.82}
+
+\lstset{ %
+  backgroundcolor=\color{white},   % choose the background color; you must add \usepackage{color} or \usepackage{xcolor}
+  basicstyle=\footnotesize,        % the size of the fonts that are used for the code
+  breakatwhitespace=false,         % sets if automatic breaks should only happen at whitespace
+  breaklines=true,                 % sets automatic line breaking
+  captionpos=b,                    % sets the caption-position to bottom
+  commentstyle=\color{mygray},    % comment style
+  deletekeywords={...},            % if you want to delete keywords from the given language
+  escapeinside={\%*}{*)},          % if you want to add LaTeX within your code
+  extendedchars=true,              % lets you use non-ASCII characters; for 8-bits encodings only, does not work with UTF-8
+  % frame=single,                    % adds a frame around the code
+  keepspaces=true,                 % keeps spaces in text, useful for keeping indentation of code (possibly needs columns=flexible)
+  keywordstyle=\color{blue},       % keyword style
+  language=Python,                 % the language of the code
+  morekeywords={*,...},            % if you want to add more keywords to the set
+  numbers=none,                    % where to put the line-numbers; possible values are (none, left, right)
+  numbersep=5pt,                   % how far the line-numbers are from the code
+  numberstyle=\tiny\color{mygray}, % the style that is used for the line-numbers
+  rulecolor=\color{black},         % if not set, the frame-color may be changed on line-breaks within not-black text (e.g. comments (green here))
+  showspaces=false,                % show spaces everywhere adding particular underscores; it overrides 'showstringspaces'
+  showstringspaces=false,          % underline spaces within strings only
+  showtabs=false,                  % show tabs within strings adding particular underscores
+  stepnumber=2,                    % the step between two line-numbers. If it's 1, each line will be numbered
+  stringstyle=\color{mygray},     % string literal style
+  tabsize=2,                       % sets default tabsize to 2 spaces
+  title=\lstname                   % show the filename of files included with \lstinputlisting; also try caption instead of title
+}
 
@@
 \DeclareRobustCommand{\orderof}{\ensuremath{\mathcal{O}}}
-\DeclareRobustCommand{\threedi}{3Di }
+\DeclareRobustCommand{\threedi}{3Di~}
 
@@
 
-% Several attempts have been made to enable the movement towards these new working methods and increased data volumes.
-% Refs Numerical (adaptive mesh): overview
-% Interactive models: IRF: ESMF, OpenMI
-% Social modelling:
-% Web based modelling:
-
 The \threedi  platform, in which these approaches and several new ones are applied, consists of a new overland flooding model (1D/2D) and a corresponding infrastructure that has been applied for interactive web-based modeling of rural and urban areas at coverages of \orderof(1000km2) at a resolution of \orderof(0.1m).
 Here we describe the advances in the numerical methods, architecture and visualization techniques that were used to create the platform.
-
-% The advances on a numerical level, by implementing a quad tree structure allowing only the relevant physics to be computed, resulting in an order of magnitude speedup have been previously presented.
-Here we present the new architecture and it's underlying components, resulting in the advances in model interaction and scalability.
-% The new multi-tier infrastructure was designed under the following conditions:
-% Every aspect of the system should be 100 times faster than the previous approach.
-% The main end user is the water-manager, who is assumed to only know how to use a browser and likes to present realistic looking results.
-% The modeling should be interactive and social, meaning that multiple people can interact with the same running model at the same time.
-% The system should be able to run both in private and public clouds. The numerical model should be able to run on a single computer on a supercomputer and in the cloud.
-% The main interface is a mobile compatible website.
-The model results are rendered on the client in a html5 web map using both bitmap and vector graphics.
-The user can interact with the model in several ways, for example by changing properties of hydraulic structures, water levels, discharge locations, rain events.
-The browser transforms these interactions into messages which are pushed back to the model through websockets.
-Model results are also pushed to the browser using websockets and on request through a Web Map Service (WMS) extended with time support.
-% The web application server forwards the websockets messages back and forth between the browser and the model.
-This application also keeps track of authorization, authentication, user preferences and the social aspects, who is watching which model, who is in control.
-% The numerical model was adapted to an interactive model by implementing the low level, non invasive BMI interface.
-This interface makes the numerical model adhere to the Hollywood principle (don't call us, we'll call you): the control of the initialize, update and finalize steps are called from outside of the numerical model.
-All variables of the model can be accessed through direct or shared memory access, allowing for communication within nanoseconds.
-The numerical model can be run as a command line executable (for operational runs), in a GUI (for schematization design), as a shared library (for data exchange) and as a message publisher/puller (for interaction).
-A separate tool generates photo-realistic renderings in 3D stereo for presentation purposes.
-The messaging publisher module, which works for any BMI compliant model, allows to push results to the browser at a rate of over 20 time per second and a latency of under 50ms.
-% A provisioning module is used to quickly deploy new updates of models schematization and to scale when more running models or WMS servers are required.
-Models can run at cloud providers and at the central server.
-Model schematizations are kept under distributed version control so that updates can be pushed and different versions of the same model can be used.
-% The new platform has been deployed at several locations in the Netherlands and in Singapore.
-In preliminary tests we found that the model has skill comparable to previous models.
-Setting the bar for performance a few order of magnitudes higher resulted in new ideas instead of improved existing ideas.
-A major part of the platform (model messaging, WMS server) has been released open source because it can also be used with other models.
-The numerical model is shared within a small community.
-% By solving the challenges new challenges arise.
-The new challenges are technical and social.
-How can we dynamically filter out the most relevant information from the platform as the bandwidth and browsers rendering power of the mobile browsers is limited and varying.
-The shift towards social and exploratory modeling and away from the modeler "in control" require new best practices.
-How do we make sure information is used correctly?
-How "true" is the information perceived?
 
@@
 On top of this core library are independent ways of addressing the core: a light-weight interface library that exposes the \ac{API}, and the original technical \ac{GUI} that acts as a sandbox for early developments.
-The core and \ac{API} offer powerful new functionality for direct and fast interaction using introspection (described in \autoref{sec:coupling} and the Hollywood-principle.
+The core and \ac{API} offer powerful new functionality for direct and fast interaction using introspection (implemented using the \ac{BMI} and the Hollywood-principle.
 
@@
 More specifically: get current water levels and flow fields, current running state of water pumps and other structures, active meteorological forcings.
-Each of these state variables can just as easily be \emph{set} as they are \emph{read}.
+Each of these state variables can just as easily be \emph{set} as they are \emph{get}.
+These functions are implemented in correspondence with the \ac{BMI} standard \citep{Peckham2013}.
 The \ac{API} exposes memory pointers to relevant model state variables.
-The model will directly 'feel' and use its new state.
+This is how the model directly `feels' and uses its new state.
 
@@
 
+\begin{enumerate}
+\item \textbf{variable = get(``name'')}: Return a variable from the model
+\item \textbf{set(``name'', variable)}: Set a variable into the model
+\end{enumerate}
+
+To reduce the number of errors a template system was introduced to generate the code used to allocate,  deallocate, and expose the variables.
+The model engine is compiled as a shared library with \emph{C} compatible external functions.
+
 The power and elegance of the subgrid concept is now also proving itself for the hydrological processes in the subsoil.
 The planter size pixel resolution of the digital terrain maps for detailed evaluation of overland flow friction and water storage is described in \citet{Stelling2012}.
-Coupling the 2D grid with 1D channels and controllable structures enables greater detail in civilized regions.
+
+To enable the steering of the water systems the 2D grid was extended 1D channels and controllable structures (for example pumps, orifices, weirs) which enables a representation of water systems in civilized regions.
 
@@
 Infiltration, evaporation and drainage are now also evaluated at pixel resolution, whereas the computational grid cells are larger and ensure high performance.
+The heterogeneous landscape and land use (be it it urban and/or rural) are specified in the model as gigapixel bitmaps for soil types, crop types, drainage resistance and maximum infiltration.
+Consider the basic head-dependent linear infiltration in (\ref{eq:qinf}), where $\zeta_\mathrm{sw}$ and $h_\mathrm{gw}$ are surface and ground water level, respectively, and $d_{i,j}$ is the surface/bed level of pixel $(i,j)$.
+Notice how dry pixels (higher land) do not amount to the total infiltration.
+This specific type of head-dependent infiltration rate can be chosen freely.
+The major advancement here is the pixel-resolution soil characteristics.
 
-The heterogeneous landscape and land use (be it it urban and/or rural) are specified in the model as gigapixel bitmaps for soil types, crop types, drainage resistance and maximum infiltration.
-Consider the basic head-dependent linear infiltration in (\ref{eq:qinf}) as an example.
-%
 \begin{equation}
@@
 \end{equation}
-where $\zeta_\mathrm{sw}$ and $h_\mathrm{gw}$ are surface and ground water level, respectively, and $d_{i,j}$ is the surface/bed level of pixel $(i,j)$.
-Notice how dry pixels (higher land) do not amount to the total infiltration.
 
-Note that this specific type of head-dependent infiltration rate in can be chosen freely, but the fact that it uses pixel-resolution soil characteristics is what is important here.
-
-%
-These advances in the numerical model allow the model to act as a component in the interactive web-based model environment that will be described in the next section.
+These advances in the numerical model allow the model to act as a component in the interactive web-based model environment with a wide variety of possible applications.
 
 \section{Architecture}
-\begin{wrapfigure}[12]{l}{0.5\textwidth}
+Now that a numerical model is available that allows high resolution interactive modelling, how can we hook it up to a tablet, mobile phone and map table, all at the same time, to allow for social and interactive modeling.
+
+To enable such an environment the most logical choice is to use a web based environment.
+Over the last years this choice has become more apparent.
+The large list of browser enhancements (for example inline \ac{SVG}, canvas, plugin free video \citep{Berjon2014} and \ac{CSS} animations \citep{Jackson2013}) allow for rich visualizations independent of the device type.
+
+An overview of the system architecture is given in \autoref{fig:arch}.
+The architectural follows the general ``Model View Controller'' pattern \citep{Gamma1994} with caching of events and data to accommodate for the model continuously running in the background.
+The implementation of this system required several technical innovations and social considerations.
+
+\begin{wrapfigure}[9]{r}{0.5\textwidth}
   \centering
-  \includegraphics[width=0.48\textwidth]{architecture}
+  % \vspace{-20pt}
+  \includegraphics[width=0.48\textwidth]{arch}
   \caption{Architecture overview}
@@
 
-To enable social and interactive modeling through a map environment the most logical choice was to use a web based environment.
-Over the last years this choice has become easier through the large list of advancements for browser based applications.
-An overview of the architecture is given in \autoref{fig:arch}.
-Notice how the architectural follows the general Model View Controller pattern \citep{GOF2004}.
-On a high level the architecture follows is classic.
-In this section we will describe the lower level, yet important,  technical innovations that made the application of this architecture successful.
-
-\begin{wrapfigure}[16]{r}{0.5\textwidth}
-  \centering
-  \includegraphics[width=0.48\textwidth]{webgui}
-  \caption{Web \ac{GUI}}
-  \label{fig:gui}
-\end{wrapfigure}
-
 The end-users, interact with the web based \ac{GUI}.
 The design philosophy is minimalistic, allowing little room for user mistakes.
-For example manholes can only be placed inside the model domain, and can be controlled by only two different discharge settings (normal and extreme discharge).
+For example manholes can only be placed inside the model domain, and can be controlled by only two different discharge settings (normal and extreme discharge) (see \autoref{fig:twod}).
 Sessions with groups of end-users resulted in the idea that this interface was the most usable.
@@
 
-Several challenges arise when preparing a model for a web-based use.
 One of the differences when comparing a desktop user to a web-based user is the short attention span.
-If no indication of progress is given for longer than two second users will start reloading, clicking erratically or will just close the website as a whole searching for a new ``frisson'' \citep{Carr2011}.
+If no indication of progress is given for longer than two second users will start reloading, clicking erratically or will just close the website, searching for a new ``frisson'' \citep{Carr2011}.
 
@@
 
+To communicate the model results the \ac{MMI} protocol is used \citep{Baart2014a}.
+This protocol is suitable for sending model variables over the internet.
+A model message consists of a metadata header in \ac{JSON} format, optionally followed by bytes.
+Listing \ref{lst:mmi} shows an example of an \ac{MMI} message.
+In this project the \ac{0MQ} protocol is used as a the underlying transport protocol.
+The advantage of using the \ac{MMI}, \ac{0MQ} combination is that it has flexible metadata, does not copy data and it is very fast.
+
+\begin{lstlisting}[caption=MMI message,label=lst:mmi]
+# The first part of the message contains metadata
+{
+  "name": "waterlevel",
+  "shape": [3,3],
+  "dtype": "float64",
+  "attributes": {
+       "standard_name": "sea_surface_altiude",
+       "units": "m"
+  }
+}
+# The second part of the message contains bytes
+\x81\xb3b\xa0\xe8 ... \x01\x90\xe5\x84\xb5\xb0
+\end{lstlisting}
+
 Interaction is another challenge where timing is crucial.
-When a user interacts with the model, for example places a discharge point, turns on a pump or lowers an orifice she expects a response within a second.
+When a user interacts with the model, for example places a discharge point, turns on a pump or lowers an orifice, the user expects a response within a second.
 This short feedback loop is achieved using several techniques.
-Using the regular \ac{HTTP} GET Request is too slow in this case.
 To reduce communication we use a chain of direct connections.
-This starts with a websocket \citep{w3c2010} connection between the message broker and the browser.
+This starts with a websocket \citep{Hickson2012} with the browser.
 The model controller gets notified directly of user interactions.
-These interactions are collected in a separate thread and pushed into the model on every timestep.
+These interactions are collected in a separate cache and pushed into the model on every timestep.
 These timesteps have to be kept shorter than in a classical model run.
@@
 
-As a final challenge there is the social aspect of modeling.
+The social aspect of modeling is another aspect where interesting problems occur.
 Several group sessions with different settings have been tried.
-The current setup is that the model is run in a channel, like in a chat room.
+The current setup is that the model is run in a ``virtual room'', like in a chat room.
 Multiple people can join the model run.
@@
 
-\subsection{Data sources }
-\begin{itemize}
-  \item lidar
-  \item rain example
-  \item structures/levees
-\end{itemize}
+\section{Visualization}
+The 3Di platform serves a wide variety of end-users.
+It is important to find a good balance between what is computed and how information is perceived.
+For scientific or engineering users one wants to provide information as close to the model results as possible.
+For less experienced end users it becomes important that the information is perceived correctly rather than that it is presented correctly.
 
-\section{Coupling}
-\label{sec:coupling}
+First a note about the performance.
+Humans will detect images at a rate of about \si{12}{\per\second} as continuous \citep{Landis1954}.
+To enable the model results to be perceived as continuous, taking into account some random network delay, the aim is to generate a map within \si{50}{\mili\second}.
+Generating results at this framerate can be done using two approaches.
+The first is to reduce the timestep so that it computes in less than \si{50}{\mili\second}.
+This is the only way to get true model results at this framerate.
+An alternative is to interpolate in time and provide the insinuation of continuous information.
+We use both approaches here.
+We reduce the calculation time so that interaction is fluent and we can pass the flicker fusion threshold.
+If one wants to use longer time steps we also visualize at a high framerate.
+An example of this is the visualization of river flow using dots that are animated with \ac{CSS} animations as a function of the current flow velocity (see \autoref{fig:oned}).
 
-\begin{itemize}
-  \item raster
-  \item quad unstructured
-  \item 1d network
-  \item structure tables
-\end{itemize}
+Another performance technique used to achieve the high framerate is to use the \emph{Mahlen nach Zahlen} method, a reference to a product from the German company Ravensburger~\textregistered.
+A map with cell numbers is prepared in advance.
+Once a quantity for a new timestep becomes available the map is filled in with the colors corresponding to the cell numbers.
+This achieves much faster rendering times than drawing the cells a geometry drawing function.
 
-\subsection{BMI}
+\begin{figure}[htbp]
+  \centering
 
-Basic Model Interface Wrapper
-The \ac{BMI} is a low level API designed access both control and data access functions of the model engine component.
-It is similar to the type introspection used in many programming languages, however, instead of the class types \ac{BMI} provides access to the model variables, including their attributes such as data type, unit of measure, input or output role.
-To provide full access to the internals of the internal model state the following function types should be implemented: model control, model information, variable information, variable getter and setter and grid information.
-The \ac{BMI} API is quite compact and does not require much effort to implement it in the model.
-However, it can be quite time consuming to continuously extend access to the model variables in programming languages such as fortran.
-To simplify this maintenance process we have implemented an automated way to add or remove model variables via \ac{BMI}.
-The automation was added using Python script and Mako template engine (Figure 1).
-
-\begin{figure}[h]
-  \centering
-  \includegraphics[width=10cm]{bmi_3di}
-  \caption{Basic Model Interface template}
-  \label{fig:bmi}
+  \begin{subfigure}{0.48\textwidth}
+    \includegraphics[width=1\textwidth]{webgui}
+    \caption{Representation of two dimensional subpixels water level}
+    \label{fig:twod}
+  \end{subfigure}
+  \begin{subfigure}{0.48\textwidth}
+    \includegraphics[width=1\textwidth]{oned}
+    \caption{Representation of one dimensional channel network using an animated stroke pattern.}
+    \label{fig:oned}
+  \end{subfigure}
+  \label{fig:gui}
+  \caption{Browser based interactive model user interface of \threedi}
 \end{figure}
 
-Figure 1: Automatic generation of the include files for \ac{BMI} support
+The variety of end-users exposes the numerical model to non-technical users with little appreciation of implemented numerical schemes.
+Some artifacts from the numerical schemes result in confusion for end users.
+The world in the subgrid model is internally represented in voxels, not unlike the computer game Minecraft.
+This results in visual artifacts as represented in \autoref{fig:arta}.
+In terrains with steeper slopes the raw model output results in a rice field like pattern.
+Another type of artifact occurs when a levee crosses a calculation cell both sides of the cell will contain water.
 
-The Mako template files used to generate FORTRAN include all code necessary to expose all required external functions of the model engine shared library.
-This library is then used by the 3Di model wrapper (developed using Python and ctypes) to control model.
+One can argue whether to show these ``true'' yet confusing results or to process the information so that is perceived as intended.
+In the \threedi platform the default visualization hides these results by filtering and smoothing the output.
+Hiding the square world is implemented using different techniques.
 
-The approach described above significantly decreased costs of change required to maintain \ac{BMI} support in the 3Di model.
+\begin{figure}[htpb]
+  \centering
+  \begin{subfigure}{0.48\textwidth}
+    \includegraphics[width=1\textwidth]{arta}
+    \caption{Uninterpolated subgrid water level}
 
-Additionally to the standard functions we had to extend BMI with support for simple compound data types (C structures using only simple fields).
-This was necessary to simplify access to the higher level model entities such as hydraulic structures (e.g.
-pumps, weirs, culverts).
+    \label{fig:arta}
+  \end{subfigure}
+  \begin{subfigure}{0.48\textwidth}
+    \includegraphics[width=1\textwidth]{artb}
+    \caption{Interpolated subgrid water level}
+    \label{fig:artb}
+  \end{subfigure}
 
-\subsection{MMI}
-\begin{itemize}
-  \item message interface
-  \item serialization
-  \item comparison to other interfaces
-\end{itemize}
-\subsection{Websockets}
-\begin{itemize}
-  \item Advantages of push for models
-  \item Serialization
-\end{itemize}
-\section{Visualization}
-\subsection{Ravensburger }
-\begin{itemize}
-  \item performance comparison
-  \item WMS/TWMS coupling
-\end{itemize}
-\subsection{Interpolation}
-Interpolation Martijn
+  \label{fig:artifacts}
 
-The 3Di platform enables a more exploratory approach to modeling, serving a wide variety of end-users.
-The variety of end-users exposes the numerical model to non-technical users with little understanding of implemented numerical schemes.
-Some artefacts resulting from the numerical schemes are therefore filtered and smoothed using an interpolation method.
-Artefacts include squared water depth visualization and water present on two sides of a levee within one calculation cell.
-The filtering and smoothing reduces the need for prior knowledge of numerical modeling.
+  \caption{Interpolation method}
 
-The numerical core of the 3Di platform is based on a subgrid calculation method (Stelling, 2012).
-The hydrodynamic calculations are performed on a non-uniform grid (quadtree ordering), with use of high resolution \ac{DEM} information as input.
-Flow variables, such as water levels and flow velocities, are available on the calculation grid, which reduces storage costs.
-The resulting adaptive storage of flow variables results in artefacts after plotting them on the high resolution \ac{DEM} rasters.
-Figure 1a shows the occurring artefacts.
-
-\begin{figure}[h]
-  \centering
-  \includegraphics[width=0.5\textwidth]{artifacts_a}
-  \caption{Interpolation method}
-  \label{fig:artifacts_a}
 \end{figure}
 
-\begin{figure}[h]
+A barycentric linear interpolation technique is selected for filtering artefacts on the subgrid level.
+This is one of the few techniques that allow for the interpolation within the rendering time constrain.
+Second, the ability to interpolate water level results on an arbitrary grid.
+The quad tree calculation grid is triangulated using Qhull \citep{Barber1996}, creating a triangulated mesh.
+Water level information is present on cell centers and is then linearly interpolated to the underlying subgrid.
+The interpolation uses a weighted average the determine water levels on the subgrid.
+Defining the triangular mesh can be done and reused with new values.
+Results of plotting water depths after the barycentric linear interpolation are shown in \autoref{fig:artb}.
+
+\begin{figure}
   \centering
-  \includegraphics[width=0.5\textwidth]{artifacts_b}
-  \caption{Interpolation method}
-  \label{fig:artifacts_b}
+  \begin{subfigure}{0.45\textwidth}
+    \includegraphics[width=1\textwidth]{levees1}
+    \caption{Redistribution of cell sections based on levee positioning.}
+    \label{fig:levee1}
+  \end{subfigure}
+  \begin{subfigure}{0.45\textwidth}
+    \includegraphics[width=1\textwidth]{levees2}
+    \caption{Interpolated subgrid water level with levee constrained}
+    \label{fig:levee2}
+  \end{subfigure}
+  \label{fig:levees}
+
+  \caption{Levee-constrained interpolation method}
+
 \end{figure}
 
-Figure 1a.  Artefacts in the visualization                      Figure 1b.  Filtered visualization
-
-A barycentric linear interpolation technique is selected for filtering artefacts on the subgrid level, because of two main restrictions.
-First, the web visualization requires a 50 ms rendering time in the browser that is still obtained using this method.
-Second, the ability to interpolate water level results on an arbitrary grid.
-The quadtree calculation grid was triangulated from the cell centers using existing software libraries, creating a triangulated mesh (Figure 2b).
-Water level information is present on cell centers and is then linearly interpolated to the underlying subgrid.
-The interpolation uses a weighted average the determine water levels on the subgrid.
-The determination of the triangular mesh is very low cost after which the interpolation can be executed immediately.
-The little calculation steps needed result in high performance.
-This technique has a large upside in calculation time compared to a technique using cell corners as interpolation points.
-When using cell corners, an extra intermediate calculation has to be made to determine water levels at cell corners.
-After the extra calculation the interpolation is executed.
-Results of plotting water depths after the barycentric linear interpolation are shown in Figure 1b.
-
-The triangulation within the linear barycentric interpolation also allows us to filter levee artefacts of the numerical model .The calculation grid incorporates the function of a levee in the numerical schemes.
+The triangulation within the linear barycentric interpolation also allows us to filter levee artefacts of the numerical model.
+The calculation grid incorporates the function of a levee in the numerical schemes.
 However it disregards the actual position of such a levee within a calculation cell.
 The triangulation can reconstruct levees with use of additional levee geometries.
-A triangle is one of the simplest geometrical forms and can be used reconstruct any arbitrary shape (Welch and Witkin, 1994).
-To incorporate the levee in the trianglulation, the levee and grid geometries are intersected (Figure 2a).
+A triangle is one of the simplest geometrical forms and can be used reconstruct any arbitrary shape \citep{Welch1994}.
+To incorporate the levee in the trianglulation, the levee and grid geometries are intersected (\autoref{fig:levee1}).
 For the intersected calculation cells the influential water level points are determined and coupled to the relevant calculation areas.
-The arrows in figure 2a indicate which water level is relevant for the dike intersected cell areas.
+The arrows in \autoref{fig:levee1} indicate which water level is relevant for the dike intersected cell areas.
 When all waterlevel points are allocated to the dike, the triangulation can be executed with the incorporation of the levee geometries.
@@
 Successively the interpolation is executed, resulting in water being retained behind the dike within one calculation cell.
-The interpolated result is shown in Figure 2b.
+The interpolated result is shown in \autoref{fig:levee2}.
 
-\begin{figure}[h]
-  \centering
-  \includegraphics[width=0.5\textwidth]{interpolationlevees3}
-  \caption{Interpolation method}
-  \label{fig:interp_a}
-\end{figure}
-
-Figure 2a. Levee and grid geometry intersections        Figure 2b. Levee incorporated interpolation
-
-Reference:
-Welch, W.
-Witkin, A.
-Free-Form Shape Design Using Triangulated Surfaces, Proceeding SIGGRAPH '94 Proceedings of the 21st annual conference on Computer graphics and interactive techniques, Pages 247-256, 1994
-Stelling, G.S. Quadtree flood simulations with sub-grid digital elevation models, Proceedings of the ICE - Water Management, Volume 165, Issue 10, 01 November 2012, pages 567 –580
-
-http://quietplease.com/home/tri.pdf
-
-\section{Limitations: \emph{All}}
-Several other models use similar approaches, albeit with limitations.
+\section{Discussion and conclusion}
+Several other models use a similar numerical approach, albeit with limitations.
 The subgrid scheme by \citet{Stelling2012} is partly based on the subgrid method for flooding and drying by \citet{Casulli2009}, who implements his methods in the UnTRIM research package \citet{Casulli2000}.
@@
 
-\section{Future plans: \emph{All}}
-In that respect, combining the subgrid approach with our own unstructured flow solver Delft3D-FLOW Flexible Mesh \citep{Kernkamp2011} which is already equipped with an interactive BMI-interface) is a logical future step.
-\section{Conclusions: \emph{Fedor}}
+The major advancement here is to redesign the numerical engine into a component that allows for external control and introspection.
+By combining this numerical model with the technical advancements described in this paper, we have enabled interactive web-based modeling at a high resolution and wide coverage.
+
+Possible future directions include combining the subgrid approach with an unstructured flow solver, such as Delft3D-FLOW Flexible Mesh \citep{Kernkamp2011}.
+The pixel based approach has the advantage that it is simple to implement and to scale.
+Unfortunately the world is not so square, especially not on larger scales, when one goes beyond the scale of a small country.
+
+Another planned step is to incorporate or preferably to couple the system with other physical models.
+The monolithic approach of putting all processes into one model has proven cumbersome in previous implementations.
+The current model controller can easily incorporate other models because the communication and control are already out of the model engine.
+Water quality, coastal and socio-economic models are possible candidates for inclusion in the platform.
+
+Viewing model results through the web is already a reality (see \citep{Blower2013} for example), many model centers around the world provide their results through readily available maps and plots.
+We have made progress to run and interact with models through the browser.
+What is not available, for complex models, is the ability to set up a new model through the web.
+At the moment it is still cumbersome to implement this functionality.
+However within the next few years it is likely that also this part of the modeling workflow will become browser-based.
 
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
+\section*{Acronyms}
 \begin{acronym}[AAAAA]
+\acro{CSS}{Cascading Style Sheets}
+\acro{MMI}{Model Message Interface}
+\acro{JSON}{JavaScript Object Notation}
+\acro{0MQ}[\O{}MQ]{Zero Message Queue}
+\acro{SVG}{Scalable Vector Graphics}
 \acro{H2O}[$\mathrm{H_2O}$]{water}

### 2014-03-21 article.tex
style matchup with Word template

@@
 \usepackage{amsmath}
-\usepackage[printonlyused]{acronym}
+\usepackage[nolist]{acronym}
 \usepackage{sansmath}
@@
 \DeclareRobustCommand{\orderof}{\ensuremath{\mathcal{O}}}
-\DeclareRobustCommand{\threedi}{3Di~}
+\DeclareRobustCommand{\threedi}{$3Di$~}
 
@@
 
-\title{Interactive web based flood modeling at country wide scale and planter size resolution.}
+\title{Interactive web-based flood modeling at country wide scale and planter size resolution.}
 
@@
   \address[A1]{\it{Deltares,
-      Rotterdamseweg
+      Rotterdamseweg,
       Delft, The Netherlands (fedor.baart@deltares.nl, arthur.vandam@deltares.nl, gennadii.donchyts@deltares.nl)}},
@@
 This model, fed with measurements, and possible measures, calculates the expected flooding.
-  The traditional modelling process consists of a three step process: schematization setup, model running and post-processing, with a total feedback time of hours.
-This process is suitable for confirmatory modelling.
+  The traditional modeling process consists of a three step process: schematization setup, model running and post-processing, with a total feedback time of hours.
+This process is suitable for confirmatory modeling.
 Most of the time, models are applied exploratory, requiring a different workflow.
-  Enabling exploratory modelling requires a shift in utilisation of the modelling instrument.
-Stakeholders are in control and together evaluate ideas by interacting with the model through a mobile compatible website, supported by the modellers’ expertise.
-Enabling this type of interactive modelling requires a new level of performance.
+  Enabling exploratory modeling requires a shift in utilisation of the modeling instrument.
+Stakeholders are in control and together evaluate ideas by interacting with the model through a mobile compatible website, supported by the modelers’ expertise.
+Enabling this type of interactive modeling requires a new level of performance.
   The \threedi platform, in which the new approach was applied, consists of a new flooding and hydrological model (1D/2D) with a corresponding cloud based infrastructure.
-Applications in rural and urban areas of \orderof(1000km2) at a resolution of \orderof(0.1m) have shown its capabilities for both exploratory and confirmatory modelling.
+Applications in rural and urban areas of \orderof(1000km2) at a resolution of \orderof(0.1m) have shown its capabilities for both exploratory and confirmatory modeling.
   The ambition that every component should be at least a 100 times faster than the previous approach, resulted in several advancements, both in the numerical engine and the software that interacts with the user and pushes the data to the web.
@@
 \begin{keyword}
+interactive, overland flooding, web-based, modeling, hydrodynamic, social modeling
 \end{keyword}
@@
 The topography and infiltration rate at a spatial resolution of a planter determine if the water will flow to ones backyard or the neighbors.
-This resulted in the question how can we design a platform that is suitable for this new modeling approach and meets the corresponding challenges.
+How can we design a platform that is suitable for this new modeling approach and meets the corresponding challenges?
 
 The \threedi  platform, in which these approaches and several new ones are applied, consists of a new overland flooding model (1D/2D) and a corresponding infrastructure that has been applied for interactive web-based modeling of rural and urban areas at coverages of \orderof(1000km2) at a resolution of \orderof(0.1m).
-Here we describe the advances in the numerical methods, architecture and visualization techniques that were used to create the platform.
+Here we describe the advances in the numerical methods, architecture and visualization techniques.
 
 \section{Numerical model}
-The computational core originates from core components of the {SOBEK} 1D2D flooding software \citep{Stelling2006}, and has undergone three major transformations. \begin{inparaenum}
-\item a streamlining of the software architecture to offer access to the memory of a running model,
+The computational core originates from components of the {SOBEK} 1D2D flooding software \citep{Stelling2006} with three major improvements: \begin{inparaenum}
+\item streamlining of the software to offer access to the memory of a running model,
 \item overland flow at subgrid scale and
-\item inclusion of hydrological processes in the subsoil also at subgrid scale.
+\item inclusion subsoil processes at subgrid scale.
 \end{inparaenum}
@@
 The traditional standalone flooding program and its integrated \ac{GUI} have been split into dedicated components.
-Most importantly, the core library with all numerical algorithms and I/O.
-On top of this core library are independent ways of addressing the core: a light-weight interface library that exposes the \ac{API}, and the original technical \ac{GUI} that acts as a sandbox for early developments.
+A core library combines all numerical algorithms.
+On top of this library a light-weight interface library exposes the \ac{API}.
+This interface is compiled as a shared library with \emph{C} compatible external functions.
 The core and \ac{API} offer powerful new functionality for direct and fast interaction using introspection (implemented using the \ac{BMI} and the Hollywood-principle.
@@
 This is how the model directly `feels' and uses its new state.
+
+\begin{enumerate}
+\item \textbf{variable = get(``name'')}: Return a variable from the model
+\item \textbf{set(``name'', variable)}: Set a variable into the model
+\end{enumerate}
 
@@
 \begin{wrapfigure}[19]{r}{0.5\textwidth}
+\vspace{-30pt}
 \centering
@@
 In between these interface calls, the model state may be changed from outside and the core should support that.
-The library is loaded and can remain a running process for a long time, for many different models..
-A positive side-effect of implementing this interface is that it may reveal subtle inconsistencies in existing code, for example when initializing different models after one another: incomplete cleanup of previous settings, already allocated memory, unclosed file pointer.
+The library is loaded and can serve as a running process for a long time.
+A positive side-effect of implementing this interface is that it reveals subtle inconsistencies in existing code, for example initializing different models after one another: incomplete cleanup of previous settings, already allocated memory, unclosed file pointer all become apparent.
+To reduce the number of errors a template system was introduced to generate the code used to allocate, deallocate, and expose the variables.
 
-\begin{enumerate}
-\item \textbf{variable = get(``name'')}: Return a variable from the model
-\item \textbf{set(``name'', variable)}: Set a variable into the model
-\end{enumerate}
+The planter size pixel resolution of the digital terrain maps for detailed evaluation of overland flow friction and water storage is described in \citet{Stelling2012}.
+To represent water systems in civilized regions the 2D grid was extended 1D channels and controllable structures (for example pumps, orifices, weirs).
+This allows for more realistic and interactive overland flooding simulation.
 
-To reduce the number of errors a template system was introduced to generate the code used to allocate,  deallocate, and expose the variables.
-The model engine is compiled as a shared library with \emph{C} compatible external functions.
+The power of the subgrid concept is also applied the hydrological processes in the subsoil.
+An important factor in potential flooding is the storage of water in the subsoil, or the lack thereof.
+The change in surface water volume is driven by several familiar processes:
+\[
+\Delta V_\mathrm{surface\ water}&=(Q_\mathrm{overland} + Q_\mathrm{precipitation} - Q_\mathrm{infiltration} - Q_\mathrm{evaporation} + Q_\mathrm{drainage})\cdot\Delta t
+\]
 
-The power and elegance of the subgrid concept is now also proving itself for the hydrological processes in the subsoil.
-The planter size pixel resolution of the digital terrain maps for detailed evaluation of overland flow friction and water storage is described in \citet{Stelling2012}.
-
-To enable the steering of the water systems the 2D grid was extended 1D channels and controllable structures (for example pumps, orifices, weirs) which enables a representation of water systems in civilized regions.
-
-Another important influence on potential flooding is the storage of water in the subsoil, or the lack thereof. The change in surface water volume is driven by several familiar processes:
-\[
-\Delta V_\mathrm{surface\ water}=(Q_\mathrm{overland} + Q_\mathrm{precipitation} - Q_\mathrm{infiltration} - Q_\mathrm{evapo(transpi)ration} + Q_\mathrm{drainage})\cdot\Delta t
-\]
-Infiltration, evaporation and drainage are now also evaluated at pixel resolution, whereas the computational grid cells are larger and ensure high performance.
-The heterogeneous landscape and land use (be it it urban and/or rural) are specified in the model as gigapixel bitmaps for soil types, crop types, drainage resistance and maximum infiltration.
-Consider the basic head-dependent linear infiltration in (\ref{eq:qinf}), where $\zeta_\mathrm{sw}$ and $h_\mathrm{gw}$ are surface and ground water level, respectively, and $d_{i,j}$ is the surface/bed level of pixel $(i,j)$.
-Notice how dry pixels (higher land) do not amount to the total infiltration.
-This specific type of head-dependent infiltration rate can be chosen freely.
-The major advancement here is the pixel-resolution soil characteristics.
-
-\begin{equation}
-\label{eq:qinf}
-Q_\mathrm{infiltration}= \sum_{\mathrm{pixels}\ i,j} \frac{\min(0, \zeta_\mathrm{sw}-\max(h_\mathrm{gw}, d_{i,j}))}{\mathrm{resistance}_{i,j}}\Delta x_p \Delta y_p,
-\end{equation}
+The heterogeneous landscape and land use are specified in the model as gigapixel bitmaps for soil types, crop types, drainage resistance and maximum infiltration.
+Infiltration, evaporation and drainage are evaluated at pixel resolution, whereas the computational grid cells are larger and ensure high performance.
 
@@
 \section{Architecture}
-Now that a numerical model is available that allows high resolution interactive modelling, how can we hook it up to a tablet, mobile phone and map table, all at the same time, to allow for social and interactive modeling.
+Now that a numerical model is available that allows high resolution interactive modeling, how can we hook it up to a tablet, mobile phone and map table, all at the same time, to allow for social and interactive modeling?
 
-To enable such an environment the most logical choice is to use a web based environment.
+To enable such an environment the most logical choice is to use a web-based environment.
 Over the last years this choice has become more apparent.
 The large list of browser enhancements (for example inline \ac{SVG}, canvas, plugin free video \citep{Berjon2014} and \ac{CSS} animations \citep{Jackson2013}) allow for rich visualizations independent of the device type.
-
 An overview of the system architecture is given in \autoref{fig:arch}.
-The architectural follows the general ``Model View Controller'' pattern \citep{Gamma1994} with caching of events and data to accommodate for the model continuously running in the background.
+The architecture follows the general ``Model View Controller'' pattern \citep{Gamma1994}, with caching of events and data to accommodate for the model continuously running in the background.
 The implementation of this system required several technical innovations and social considerations.
@@
   \centering
-  % \vspace{-20pt}
+  \vspace{-20pt}
   \includegraphics[width=0.48\textwidth]{arch}
@@
 
-The end-users, interact with the web based \ac{GUI}.
+The end-users, interact with the web-based \ac{GUI}.
 The design philosophy is minimalistic, allowing little room for user mistakes.
-For example manholes can only be placed inside the model domain, and can be controlled by only two different discharge settings (normal and extreme discharge) (see \autoref{fig:twod}).
-Sessions with groups of end-users resulted in the idea that this interface was the most usable.
-The design goals of providing freedom to explore together with high resolution high frequency model output, yet without the insinuation that the model results are ``true'' proves to be challenging.
-Many end-users confuse high resolution, detailed tuning parameters and higher frequency results with higher accuracy.
+For example discharge locations only have two options (normal and extreme discharge) (see \autoref{fig:twod}).
+Sessions with groups of end-users resulted in this design that provides freedom to explore without the insinuation that the model results are ``true''.
 
@@
 If no indication of progress is given for longer than two second users will start reloading, clicking erratically or will just close the website, searching for a new ``frisson'' \citep{Carr2011}.
-
-To make sure that the model is ready to run when a user arrives and no time is lost with starting up, initializing, reading files a lot of attention is given to the provisioning of the model.
-Provisioning in this context is preparing a model controller so that it starts it's computation on request.
+To make sure that the model is ready to run when a user arrives a lot of attention is given to the provisioning of the model.
+Provisioning, in this context, is preparing a model controller so that it starts its computation on request.
 Machines are prepared and put on hold at one of the \ac{IaaS} providers (such as Amazon).
 These machines are prepared with a container that contains the model schematization, tables and grids and hot restart files.
-The tables, grids and restart files are read into memory so that the model is ready to start its computation when it is needed.
-This results in a time between a user logging in and a model start below 5 seconds.
+These files are read into memory so that the model is ready to start its computation when needed, resulting in a time between a user logging in and a model start in under 5 seconds.
 
@@
 In this project the \ac{0MQ} protocol is used as a the underlying transport protocol.
-The advantage of using the \ac{MMI}, \ac{0MQ} combination is that it has flexible metadata, does not copy data and it is very fast.
+The advantage of using the \ac{MMI}, \ac{0MQ} combination is that it has flexible metadata, does not copy data is very fast.
 
@@
 This short feedback loop is achieved using several techniques.
-To reduce communication we use a chain of direct connections.
-This starts with a websocket \citep{Hickson2012} with the browser.
-The model controller gets notified directly of user interactions.
-These interactions are collected in a separate cache and pushed into the model on every timestep.
-These timesteps have to be kept shorter than in a classical model run.
-A timestep resulting in a computation time of \si{100}{\mili\second} is common.
+User interactions are directly send through websockets \citep{Hickson2012} to the model controller.
+These interactions are cached and pushed into the model on every timestep.
+These timesteps have to be kept shorter than in a classical model run, \si{100}{\mili\second} is common.
 After a model timestep the chain is inverted.
-The model pushes it's arrays back in the direction of the browser, which is subscribed to these events and can directly visualize the results.
-This message broker approach allows interactions with a feedback time in the order of \si{100}{\mili\second}, which gives a good sense of interactivity.
+The model pushes it's arrays back in the direction of the browser, which is subscribed to these messages and can directly visualize the results.
+This fast feedback loop gives a good sense of interactivity.
 
@@
 Several group sessions with different settings have been tried.
-The current setup is that the model is run in a ``virtual room'', like in a chat room.
+The current setup is that the model is run in a ``virtual room'', like a chat room.
 Multiple people can join the model run.
@@
 Other users can request control of the model.
-The sessions worked best when a topic expert, not necessary a modeler was present in the session.
+The sessions worked best when a topic expert, not necessary a modeler, was present in the session.
 
 \section{Visualization}
-The 3Di platform serves a wide variety of end-users.
+The \threedi platform serves a wide variety of end-users.
 It is important to find a good balance between what is computed and how information is perceived.
@@
 Once a quantity for a new timestep becomes available the map is filled in with the colors corresponding to the cell numbers.
-This achieves much faster rendering times than drawing the cells a geometry drawing function.
+This achieves much faster rendering times than drawing the cells using geometry drawing function.
 
@@
 
-  \begin{subfigure}{0.48\textwidth}
+  \begin{subfigure}{0.4\textwidth}
     \includegraphics[width=1\textwidth]{webgui}
@@
   \end{subfigure}
-  \begin{subfigure}{0.48\textwidth}
+  \begin{subfigure}{0.4\textwidth}
     \includegraphics[width=1\textwidth]{oned}
@@
 The variety of end-users exposes the numerical model to non-technical users with little appreciation of implemented numerical schemes.
-Some artifacts from the numerical schemes result in confusion for end users.
+Some artifacts from the numerical schemes result in confusion.
 The world in the subgrid model is internally represented in voxels, not unlike the computer game Minecraft.
-This results in visual artifacts as represented in \autoref{fig:arta}.
-In terrains with steeper slopes the raw model output results in a rice field like pattern.
+This results in visual artifacts, as represented in \autoref{fig:arta}.
 Another type of artifact occurs when a levee crosses a calculation cell both sides of the cell will contain water.
@@
 One can argue whether to show these ``true'' yet confusing results or to process the information so that is perceived as intended.
-In the \threedi platform the default visualization hides these results by filtering and smoothing the output.
+In the \threedi platform the default visualization hides these artifacts by filtering and smoothing the output.
 Hiding the square world is implemented using different techniques.
@@
   \centering
-  \begin{subfigure}{0.48\textwidth}
+  \begin{subfigure}{0.4\textwidth}
     \includegraphics[width=1\textwidth]{arta}
     \caption{Uninterpolated subgrid water level}
-
     \label{fig:arta}
   \end{subfigure}
-  \begin{subfigure}{0.48\textwidth}
+  \begin{subfigure}{0.4\textwidth}
     \includegraphics[width=1\textwidth]{artb}
@@
   \end{subfigure}
-
   \label{fig:artifacts}
-
   \caption{Interpolation method}
-
 \end{figure}
 
-A barycentric linear interpolation technique is selected for filtering artefacts on the subgrid level.
-This is one of the few techniques that allow for the interpolation within the rendering time constrain.
-Second, the ability to interpolate water level results on an arbitrary grid.
-The quad tree calculation grid is triangulated using Qhull \citep{Barber1996}, creating a triangulated mesh.
-Water level information is present on cell centers and is then linearly interpolated to the underlying subgrid.
-The interpolation uses a weighted average the determine water levels on the subgrid.
-Defining the triangular mesh can be done and reused with new values.
-Results of plotting water depths after the barycentric linear interpolation are shown in \autoref{fig:artb}.
+A barycentric linear interpolation technique is used.
+This is one of the few techniques that allow for the interpolation within the rendering time constraint.
+It also allows to interpolate water level results on an arbitrary grid.
+The quad tree grid is triangulated using Qhull \citep{Barber1996}.
+Water level information, from the cell centers and is linearly interpolated to the underlying subgrid.
+The interpolation uses a weighted average the determine water levels.
+The triangular mesh can be reused.
+\autoref{fig:artb} shows the results of this technique .
 
-\begin{figure}
+\begin{figure}[htbp]
   \centering
-  \begin{subfigure}{0.45\textwidth}
+  \begin{subfigure}{0.4\textwidth}
     \includegraphics[width=1\textwidth]{levees1}
@@
   \end{subfigure}
-  \begin{subfigure}{0.45\textwidth}
+  \begin{subfigure}{0.4\textwidth}
     \includegraphics[width=1\textwidth]{levees2}
@@
 
+A lot of technical and numerical problems have been solved, but the methodological challenges of the new working methods are proving more difficult.
+An open issue is that end-users tend to confuse high resolution and higher frequency results with a high accuracy, a topic for further research.
+
 Possible future directions include combining the subgrid approach with an unstructured flow solver, such as Delft3D-FLOW Flexible Mesh \citep{Kernkamp2011}.
@@
 
-Viewing model results through the web is already a reality (see \citep{Blower2013} for example), many model centers around the world provide their results through readily available maps and plots.
+Viewing model results through the web is already a reality (see \citep{Blower2013} for example).
 We have made progress to run and interact with models through the browser.
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section*{Acronyms}
 \begin{acronym}[AAAAA]

### 2014-05-15 article.tex
remove dos endings, fix style

@@
 \documentclass[a4paper]{article}
+
 \usepackage{iemss}
@@
 % Don't color hyperlinks
-\usepackage{siunitx}
+
 \usepackage{wrapfig}
@@
 
+\usepackage{siunitx}
 % \RequirePackage{lineno} % for line numbers
@@
 \[
-\Delta V_\mathrm{surface\ water}&=(Q_\mathrm{overland} + Q_\mathrm{precipitation} - Q_\mathrm{infiltration} - Q_\mathrm{evaporation} + Q_\mathrm{drainage})\cdot\Delta t
+\Delta V_\mathrm{surface\ water} = (Q_\mathrm{overland} + Q_\mathrm{precipitation} - Q_\mathrm{infiltration} - Q_\mathrm{evaporation} + Q_\mathrm{drainage})\cdot\Delta t
 \]
@@
 These interactions are cached and pushed into the model on every timestep.
-These timesteps have to be kept shorter than in a classical model run, \si{100}{\mili\second} is common.
+These timesteps have to be kept shorter than in a classical model run, \SI{100}{\milli\second} is common.
 After a model timestep the chain is inverted.
@@
 First a note about the performance.
-Humans will detect images at a rate of about \si{12}{\per\second} as continuous \citep{Landis1954}.
-To enable the model results to be perceived as continuous, taking into account some random network delay, the aim is to generate a map within \si{50}{\mili\second}.
+Humans will detect images at a rate of about \SI{12}{\per\second} as continuous \citep{Landis1954}.
+To enable the model results to be perceived as continuous, taking into account some random network delay, the aim is to generate a map within \SI{50}{\milli\second}.
 Generating results at this framerate can be done using two approaches.
-The first is to reduce the timestep so that it computes in less than \si{50}{\mili\second}.
+The first is to reduce the timestep so that it computes in less than \SI{50}{\milli\second}.
 This is the only way to get true model results at this framerate.
@@
     \includegraphics[width=1\textwidth]{oned}
+
     \caption{Representation of one dimensional channel network using an animated stroke pattern.}
@@
 \bibliographystyle{iemss}
-\bibliography{bibliography}
+\bibliography{./bibliography}
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

### 2014-05-16 article.tex
fix reviewer 1 and fix header layout to line up with word based pdf

@@
 % Don't color hyperlinks
+\pdfpagewidth=210 true mm
+\pdfpageheight=297 true mm
 
@@
 \usepackage[draft]{hyperref}
+\usepackage{xcolor}
+% \hypersetup{
+%   colorlinks=true,
+%   citecolor=black,
+%   filecolor=black,
+%   linkcolor=black,
+%   urlcolor=blue
+% }
+\urlstyle{same}
 \usepackage{amsmath}
@@
 }
-
-\pdfpagewidth=210 true mm
-\pdfpageheight=297 true mm
 
@@
 The flooding of rural and urban areas is an increasing hazard to society.
-Accurate and timely predictions are essential for the water manager to be able to carefully plan, design and control the effects of flooding hazard.
+Accurate and timely predictions are essential for the water manager to be able to carefully plan, design, and control the effects of flooding hazard.
 For the prediction of these flooding hazards requires a numerical model that represents the physical processes (rain, evaporation, flow, infiltration).
 This model is fed with measurements from current or historic events.
-The results of these calculations, the expected flooding, is used for damage estimates, dike design, spatial and emergency planning and for coordinating response efforts.
+The results of these calculations, the expected flooding, is used for damage estimates, dike design, spatial and emergency planning, and for coordinating response efforts.
 Several challenges required an innovative redesign of both the numerical model and the information infrastructure where it is nested within.
@@
 There is an increase in supply of data caused by the ever growing measurement resolution and spatial coverage.
-For example the topology of waterbodies at a spatial coverage of a small country are used to determine if the water will find its way into ones neighborhood.
-The topography and infiltration rate at a spatial resolution of a planter determine if the water will flow to ones backyard or the neighbors.
+For example, the topology of waterbodies at a spatial coverage of a small country is used to determine if the water will find its way into one's neighborhood.
+The topography and infiltration rate at a spatial resolution of a planter determine if the water will flow to one's backyard or the neighbors.
 How can we design a platform that is suitable for this new modeling approach and meets the corresponding challenges?
@@
 The \threedi  platform, in which these approaches and several new ones are applied, consists of a new overland flooding model (1D/2D) and a corresponding infrastructure that has been applied for interactive web-based modeling of rural and urban areas at coverages of \orderof(1000km2) at a resolution of \orderof(0.1m).
-Here we describe the advances in the numerical methods, architecture and visualization techniques.
+Here we describe the advances in the numerical methods, architecture, and visualization techniques.
 
@@
 By introspection, the outside world (e.g., a Python-based website, a C\#-\ac{GUI}, a MATLAB-script) can examine model state at run time.
-More specifically: get current water levels and flow fields, current running state of water pumps and other structures, active meteorological forcings.
+More specifically: get current water levels and flow fields, current running state of water pumps, and other structures, active meteorological forcings.
+% Fix sentence
  Each of these state variables can just as easily be \emph{set} as they are \emph{get}.
@@
 
-The heterogeneous landscape and land use are specified in the model as gigapixel bitmaps for soil types, crop types, drainage resistance and maximum infiltration.
-Infiltration, evaporation and drainage are evaluated at pixel resolution, whereas the computational grid cells are larger and ensure high performance.
+The heterogeneous landscape and land use are specified in the model as gigapixel bitmaps for soil types, crop types, drainage resistance, and maximum infiltration.
+Infiltration, evaporation, and drainage are evaluated at pixel resolution, whereas the computational grid cells are larger and ensure high performance.
 
@@
 Over the last years this choice has become more apparent.
-The large list of browser enhancements (for example inline \ac{SVG}, canvas, plugin free video \citep{Berjon2014} and \ac{CSS} animations \citep{Jackson2013}) allow for rich visualizations independent of the device type.
+The large list of browser enhancements (for example inline \ac{SVG}, canvas, plugin free video \citep{Berjon2014}, and \ac{CSS} animations \citep{Jackson2013}) allow for rich visualizations independent of the device type.
 An overview of the system architecture is given in \autoref{fig:arch}.
@@
 Machines are prepared and put on hold at one of the \ac{IaaS} providers (such as Amazon).
-These machines are prepared with a container that contains the model schematization, tables and grids and hot restart files.
+These machines are prepared with a container that contains the model schematization, tables and grids, and hot restart files.
 These files are read into memory so that the model is ready to start its computation when needed, resulting in a time between a user logging in and a model start in under 5 seconds.
@@
 
-Another performance technique used to achieve the high framerate is to use the \emph{Mahlen nach Zahlen} method, a reference to a product from the German company Ravensburger~\textregistered.
+Another performance technique used to achieve the high framerate is to use the ``Mahlen nach Zahlen'' method, a reference to a product from the German company Ravensburger~\textregistered.
 A map with cell numbers is prepared in advance.
@@
 The current model controller can easily incorporate other models because the communication and control are already out of the model engine.
-Water quality, coastal and socio-economic models are possible candidates for inclusion in the platform.
+Water quality, coastal, and socio-economic models are possible candidates for inclusion in the platform.
 
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\bibliographystyle{iemss}
+\bibliographystyle{apalike}
 \bibliography{./bibliography}

### 2014-05-16 review/rebuttal.txt
textual changes, spelling

## Reviewer 1
> Review:	Check for comma rules on all enumerations througout the paper,
> put a comma before the last "and".

serial commas used, it's/its fixed

> Page 3, para 3:
> "determines", "one's"

determines-> determine
one's ones used correctly

> page 5, para 4:
> put units after "12" and "50" (fps ?)
Correct latex macro \si -> \SI used, fixes units throughout the paper

> page 5, para 5:
> "Malen nach Zahlen"

Changed \emph to ``''

## Reviewer 2
> Review:	This paper describes the 3Di platform, a new overland flooding and hydrological model. The paper primarily discusses the model's software structure (authors call this "numerical model), architecture, and visualization techniques. The paper is divided into 3 primary sections corresponding to these categories.
> 1. Typo - Section 2, Paragraph 7:
> "To represent water systems in civilized regions the 2D grid was extended <<insert:with>> 1D channels"

Changed

> 2. Typo - S2, para 8:
> "the subgrid concept is also applied <<insert:to>> the hydrological processes in the subsoil"

Changed

> 3. Typo - S2, para 9: oxford comma - drainage resistance, and maximum infiltration

Serial commas applied consistently

> 4. Missing discussion - S3, Para 1:
> The author's answers to "how can we hook it up to a tablet, mobile phone and map table, all at the same time, to allow for social and interactive modeling", is to "use the web" based on the corresponding text. How is the web based approach adequate for tablets and mobile phones? There is no discussion regarding the use of these devices.

Added in system description and in discussion.

> 5. Typo - S3, para 4: "If no indication of progress is given for longer than two second<<insert:s>> users will start reloading"

Units are fixed throughout the paper

> 6. Bad reference - The reference to Baart 2014 is incomplete. There is no document, article or wbesite described. How does one access this content?
DOI added to sourcecode of the python implementation.

> 7. Confusion - What is the purpose of the "bytes" following the JSON in a model message? This is not described at all.
Bytes replaced by serialized array.

> 8. Typo - S3, para 5 "The advantage of using the MMI, 0MQ combination is that it has flexible metadata, does not copy data, <<insert:and>> is very fast.

Changed.

> 9. Typo - S3, para 6 "User interactions are directly <<change to:sent>> through websockets
Changed.

> 10. Typo - S3, para 6, "After a model timestep <<insert:,>> the chain is inverted.

Changed

> 11. Typo - S3, para 6, "This fast feedback loop <<change to:supports user interactivity with the model.>>

Changed

> 12. Revise - S3, para 7, "The social aspect of modeling is another aspect"

The social aspect is another source of interesting problems.

> 13. Clarify - S4, para 2, What are the units for the numbers in this paragraph? Are the milliseconds? seconds? The reader is left to infer the units. I guess they are probably milliseconds.

Units fixed throughout the paper

> 14. Explain - "The first is to reduce the timestemp so that it computes in less that 50."
> Are you referring to the modeling timestep? Please clatify.

Changed to modeling timestep

> 15. S3, para 4 Define what a voxel is.

Defined as a picture element (pixel) with volume.

> 16. Typo - S3, para 6 "One can argue whether to show these 'true' yet confusing results or to process the information so that <<insert:it>> is perceived as intended."

Changed

> 17. Figure 4 & 5 captions - add the interpolation method type, e.g. barycentric...

Changed

> 18. Typo - S3, para 7, Revise third sentence to "Is also allows water level results to be interpolated."

Changed to It...

> 19. Typo - S3, para 7, "from the cell centers <<delete:and>> is linearly interpolated

Changed

> 20. Typo - S3, para 7, "uses a weighted average <<change:to>> determine water levels."

Changed

> 21. Revise - S3, para 8, "<<delete:The>> Triangulation with the barycentric <<moved:linear>> interpolation also allow us to

The triangular mesh can be reused to filter levee artefacts. The triangulation can reconstruct levees with use of additional levee geometries.

> 22. The conclusion is weak.

> 23. Sections 2 and 3 seem to overlap. Both contain various aspects regaarding the software framework and architecture. The paper has not been cleanly organized.

> 24. There is no discussion regarding the accuracy of the flood model itself. The paper primarily presents the software structure, features, and visualization capabilities.

> 25. The authors briefly mention how 3Di supports collaboration. It would be nice to see more discussion.

> 26. This paper appears to discuss both the 3Di modeling platform/framework and a new flooding/hydrological model. Does the flood model have a name?

> 27. Typo: Section 1, After subtitles add ":"'s, for example "Mono-disciplinary versus integrated <<insert ":">>"

Changed

> 28. The discussion and relevance to cloud based modeling is very limited in this paper. There is only a brief mention of "IaaS" and "Amazon". It is unclear what has been tested and run using cloud computing techniques.

> 29. There is no discussion from the abstract on how or if 3Di achieves "100 times faster than the previous approach". What is the previous approach? What performance evaluations have been conducted?

### 2014-05-16 article.tex
textual changes, spelling

@@
 \begin{itemize}
-\item \emph{Mono-disciplinary versus integrated} Models used to be created by one or few people working in the same discipline.
+\item \emph{Mono-disciplinary versus integrated:} Models used to be created by one or few people working in the same discipline.
 The challenge now is to model the integrated environment, which can best be approached as a community \citep{Voinov2010} of different disciplines.
-\item \emph{Confirmatory versus exploratory} From assuming that the model is valid and can be used for hypothesis testing to realizing it's a made-up simplification of reality \citep{Oreskes1994} that can be used to explore possibilities.
-\item \emph{Passive versus interactive} Moving from the pre-processing, wait a few hours, post-processing cycle towards a millisecond feedback loop with high resolution and connected physical processes.
+\item \emph{Confirmatory versus exploratory:} From assuming that the model is valid and can be used for hypothesis testing to realizing it's a made-up simplification of reality \citep{Oreskes1994} that can be used to explore possibilities.
+\item \emph{Passive versus interactive:} Moving from the pre-processing, wait a few hours, post-processing cycle towards a millisecond feedback loop with high resolution and connected physical processes.
 The challenge here is to move from the scale of a square kilometer \citep[for example][]{Losasso2008} to the scale of small country.
-\item \emph{Solitary versus social} The solitary modeler creating static pictures for a 600 page report describing all the different scenarios to a group of stakeholders in control \footnote{preferably with guidance from domain experts} of a model running simultaneous on their mobile devices and a big screen in the control room.
+\item \emph{Solitary versus social:} The solitary modeler creating static pictures for a 600 page report describing all the different scenarios to a group of stakeholders in control \footnote{preferably with guidance from domain experts} of a model running simultaneous on their mobile devices and a big screen in the control room.
 \end{itemize}
@@
 The planter size pixel resolution of the digital terrain maps for detailed evaluation of overland flow friction and water storage is described in \citet{Stelling2012}.
-To represent water systems in civilized regions the 2D grid was extended 1D channels and controllable structures (for example pumps, orifices, weirs).
+To represent water systems in civilized regions the 2D grid was extended with 1D channels and controllable structures (for example pumps, orifices, weirs).
 This allows for more realistic and interactive overland flooding simulation.
 
-The power of the subgrid concept is also applied the hydrological processes in the subsoil.
+The power of the subgrid concept is also applied to the hydrological processes in the subsoil.
 An important factor in potential flooding is the storage of water in the subsoil, or the lack thereof.
@@
 Sessions with groups of end-users resulted in this design that provides freedom to explore without the insinuation that the model results are ``true''.
+The application of \ac{CSS} and \ac{SVG} allows the interface to scale between different screen resolutions so that it can be used on different devices (phones, tablets, browsers, map table).
 
@@
 This protocol is suitable for sending model variables over the internet.
-A model message consists of a metadata header in \ac{JSON} format, optionally followed by bytes.
+A model message consists of a metadata header in \ac{JSON} format, optionally followed by the serialized array.
 Listing \ref{lst:mmi} shows an example of an \ac{MMI} message.
 In this project the \ac{0MQ} protocol is used as a the underlying transport protocol.
-The advantage of using the \ac{MMI}, \ac{0MQ} combination is that it has flexible metadata, does not copy data is very fast.
+The advantage of using the \ac{MMI}, \ac{0MQ} combination is that it has flexible metadata, does not copy data, and is very fast.
 
@@
 }
-# The second part of the message contains bytes
+# The second part of the message contains a serialized array
 \x81\xb3b\xa0\xe8 ... \x01\x90\xe5\x84\xb5\xb0
@@
 This short feedback loop is achieved using several techniques.
-User interactions are directly send through websockets \citep{Hickson2012} to the model controller.
+User interactions are directly sent through websockets \citep{Hickson2012} to the model controller.
 These interactions are cached and pushed into the model on every timestep.
 These timesteps have to be kept shorter than in a classical model run, \SI{100}{\milli\second} is common.
-After a model timestep the chain is inverted.
-The model pushes it's arrays back in the direction of the browser, which is subscribed to these messages and can directly visualize the results.
-This fast feedback loop gives a good sense of interactivity.
+After a model timestep, the chain is inverted.
+The model pushes its arrays back in the direction of the browser, which is subscribed to these messages and can directly visualize the results.
+This fast feedback loop supports user interactivity with the model.
 
-The social aspect of modeling is another aspect where interesting problems occur.
+The social aspect is another source of interesting problems.
 Several group sessions with different settings have been tried.
@@
 Generating results at this framerate can be done using two approaches.
-The first is to reduce the timestep so that it computes in less than \SI{50}{\milli\second}.
+The first is to reduce the modeling timestep so that it computes in less than \SI{50}{\milli\second}.
 This is the only way to get true model results at this framerate.
@@
 Some artifacts from the numerical schemes result in confusion.
-The world in the subgrid model is internally represented in voxels, not unlike the computer game Minecraft.
+The world in the subgrid model is internally represented in voxels (a picture element (pixel) with volume), not unlike the computer game Minecraft.
 This results in visual artifacts, as represented in \autoref{fig:arta}.
@@
 
-One can argue whether to show these ``true'' yet confusing results or to process the information so that is perceived as intended.
+One can argue whether to show these ``true'' yet confusing results or to process the information so that it is perceived as intended.
 In the \threedi platform the default visualization hides these artifacts by filtering and smoothing the output.
@@
   \label{fig:artifacts}
-  \caption{Interpolation method}
+  \caption{Linear barycentric interpolation method}
 \end{figure}
@@
 This is one of the few techniques that allow for the interpolation within the rendering time constraint.
-It also allows to interpolate water level results on an arbitrary grid.
+It also allows water level results to be interpolated.
 The quad tree grid is triangulated using Qhull \citep{Barber1996}.
-Water level information, from the cell centers and is linearly interpolated to the underlying subgrid.
-The interpolation uses a weighted average the determine water levels.
-The triangular mesh can be reused.
+Water level information, from the cell centers is linearly interpolated to the underlying subgrid.
+The interpolation uses a weighted average to determine water levels.
 \autoref{fig:artb} shows the results of this technique .
@@
     \includegraphics[width=1\textwidth]{levees2}
-    \caption{Interpolated subgrid water level with levee constrained}
+    \caption{Interpolated subgrid water level with levee constraint}
     \label{fig:levee2}
@@
 
-The triangulation within the linear barycentric interpolation also allows us to filter levee artefacts of the numerical model.
-The calculation grid incorporates the function of a levee in the numerical schemes.
-However it disregards the actual position of such a levee within a calculation cell.
+The triangular mesh can be reused to filter levee artefacts.
 The triangulation can reconstruct levees with use of additional levee geometries.
-A triangle is one of the simplest geometrical forms and can be used reconstruct any arbitrary shape \citep{Welch1994}.
+A triangle is the simplest geometrical form and can be used reconstruct any arbitrary shape \citep{Welch1994}.
 To incorporate the levee in the trianglulation, the levee and grid geometries are intersected (\autoref{fig:levee1}).
@@
 By combining this numerical model with the technical advancements described in this paper, we have enabled interactive web-based modeling at a high resolution and wide coverage.
+The use of modern web technologies allows the same interface to be used on a browser, a phone and on a map table.
 

### 2014-05-18 article.tex
writing

@@
 
-The \threedi  platform, in which these approaches and several new ones are applied, consists of a new overland flooding model (1D/2D) and a corresponding infrastructure that has been applied for interactive web-based modeling of rural and urban areas at coverages of \orderof(1000km2) at a resolution of \orderof(0.1m).
-Here we describe the advances in the numerical methods, architecture, and visualization techniques.
+The \threedi  platform, in which these approaches and several new ones are applied, consists of a new overland flooding model (1D/2D) and a corresponding infrastructure.
+This platform enables interactive web-based modeling of rural and urban areas at coverages of \orderof(1000km2) at a resolution of \orderof(0.1m).
+Here we describe the advances in the numerical model, architecture, and visualization techniques.
 
@@
 \item overland flow at subgrid scale and
-\item inclusion subsoil processes at subgrid scale.
-\end{inparaenum}
+\item inclusion subsoil processes at subgrid scale
+\end{inparaenum}.
 
@@
 The library is loaded and can serve as a running process for a long time.
-A positive side-effect of implementing this interface is that it reveals subtle inconsistencies in existing code, for example initializing different models after one another: incomplete cleanup of previous settings, already allocated memory, unclosed file pointer all become apparent.
+A positive side-effect of implementing this interface is that it reveals subtle inconsistencies in existing code, for example initializing different models after one another: incomplete cleanup of previous settings, already allocated memory, and unclosed file pointer all become apparent.
 To reduce the number of errors a template system was introduced to generate the code used to allocate, deallocate, and expose the variables.
 
-The planter size pixel resolution of the digital terrain maps for detailed evaluation of overland flow friction and water storage is described in \citet{Stelling2012}.
+Detailed evaluation of overland flow friction and water storage are simulated at subgrid scale.
+This enables to compute flooding at the planter size (\SI{0.5}{\meter}) pixel resolution of digital terrain maps.
+It also enables a speedup of a factor 100, as described in \citet{Stelling2012}.
 To represent water systems in civilized regions the 2D grid was extended with 1D channels and controllable structures (for example pumps, orifices, weirs).
@@
 
-These advances in the numerical model allow the model to act as a component in the interactive web-based model environment with a wide variety of possible applications.
+These advantages allow the model to act as a component in the interactive web-based model environment with a wide variety of possible applications.
 
@@
 \section{Discussion and conclusion}
+
+% The major finding
+The major advancement is that we have enabled interactive web-based modeling at a high resolution and wide coverage.
+The interactive, exploratory, overland flooding simulations can be used in sessions with policy makers to help them make sensible decisions. Several developments have enabled this advancement.
+
+% Interactive
+The redesign of the numerical engine follows the trends set in inter-model coupling. Applying the hollywood principle transformed the engine from a static executable into an interactive component.
+
+% Web based
+Viewing model results through the web was already a reality (see \citep{Blower2013} for example).
+We have made progress to run and interact with models through the browser.
+The use of modern web technologies allows the same interface to be used on a browser, a phone and on a map table.
+What is not available, for complex models, is the ability to set up a new model through the web.
+At the moment it is still cumbersome to implement this functionality.
+However within the next few years it is likely that also this part of the modeling workflow will become browser-based.
+
+% High resolution
+The speedup of the model by applying the subgrid method allows for a faster interaction time and a more detailed resolution.
+% But we were not the first
 Several other models use a similar numerical approach, albeit with limitations.
@@
 
-The major advancement here is to redesign the numerical engine into a component that allows for external control and introspection.
-By combining this numerical model with the technical advancements described in this paper, we have enabled interactive web-based modeling at a high resolution and wide coverage.
-The use of modern web technologies allows the same interface to be used on a browser, a phone and on a map table.
+% Limitations
+The \threedi platform is suitable for exploratory purposes.
+Work has started on a skillbed for the \threedi platform.
+A skillbed is an automated test suite that compares results to analytical cases, measurements, and results from other numerical models.
 
+% Future research
 A lot of technical and numerical problems have been solved, but the methodological challenges of the new working methods are proving more difficult.
@@
 
+% the world is not square
 Possible future directions include combining the subgrid approach with an unstructured flow solver, such as Delft3D-FLOW Flexible Mesh \citep{Kernkamp2011}.
@@
 
+% don't put everything into 1 model
 Another planned step is to incorporate or preferably to couple the system with other physical models.
@@
 Water quality, coastal, and socio-economic models are possible candidates for inclusion in the platform.
-
-Viewing model results through the web is already a reality (see \citep{Blower2013} for example).
-We have made progress to run and interact with models through the browser.
-What is not available, for complex models, is the ability to set up a new model through the web.
-At the moment it is still cumbersome to implement this functionality.
-However within the next few years it is likely that also this part of the modeling workflow will become browser-based.
 

### 2014-05-19 review/rebuttal.txt
writing

@@
 
+Rewritten
+
 > 23. Sections 2 and 3 seem to overlap. Both contain various aspects regaarding the software framework and architecture. The paper has not been cleanly organized.
+
+Reformulated title so that the content is more clear. Removed some content.
 
@@
 
+Put note that it is suitable for exploratory purposes and that validation is ongoing. Reference to confidence checklist added.
+
 > 25. The authors briefly mention how 3Di supports collaboration. It would be nice to see more discussion.
 
+Reference added to paper that discusses this topic in detail.
+
 > 26. This paper appears to discuss both the 3Di modeling platform/framework and a new flooding/hydrological model. Does the flood model have a name?
+
+The model does not have a name but is now referenced to as the 3Di computational core.
 
@@
 
+Explained in more detail
+
 > 29. There is no discussion from the abstract on how or if 3Di achieves "100 times faster than the previous approach". What is the previous approach? What performance evaluations have been conducted?
+
+Put reference to Stelling for numerical part. The visualization part was also in this order of magnitude faster, but I have no time to create reproducable numbers.
+The introduction discusses this as an ambition.
+
+Furthermore:
+General cleanup, reduce the use of the word model. Remove redundant sentences.

### 2014-05-19 article.tex
writing

@@
 This model, fed with measurements, and possible measures, calculates the expected flooding.
-  The traditional modeling process consists of a three step process: schematization setup, model running and post-processing, with a total feedback time of hours.
+  The traditional working method consists of a three step process: schematization setup, running and post-processing, with a total feedback time of hours.
 This process is suitable for confirmatory modeling.
 Most of the time, models are applied exploratory, requiring a different workflow.
-  Enabling exploratory modeling requires a shift in utilisation of the modeling instrument.
+  Enabling exploratory modeling requires a shift in utilisation of the instrument.
 Stakeholders are in control and together evaluate ideas by interacting with the model through a mobile compatible website, supported by the modelers’ expertise.
-Enabling this type of interactive modeling requires a new level of performance.
+Enabling this type of interactivity requires a new level of performance.
   The \threedi platform, in which the new approach was applied, consists of a new flooding and hydrological model (1D/2D) with a corresponding cloud based infrastructure.
@@
 
-\section{Numerical model}
-The computational core originates from components of the {SOBEK} 1D2D flooding software \citep{Stelling2006} with three major improvements: \begin{inparaenum}
+\section{Numerical model: from passive to interactive}
+The \threedi computational core originates from components of the {SOBEK} 1D2D flooding software \citep{Stelling2006} with three major improvements: \begin{inparaenum}
 \item streamlining of the software to offer access to the memory of a running model,
@@
 
-The traditional standalone flooding program and its integrated \ac{GUI} have been split into dedicated components.
-A core library combines all numerical algorithms.
-On top of this library a light-weight interface library exposes the \ac{API}.
-This interface is compiled as a shared library with \emph{C} compatible external functions.
-The core and \ac{API} offer powerful new functionality for direct and fast interaction using introspection (implemented using the \ac{BMI} and the Hollywood-principle.
+To enable interactive use, the flooding software was stripped of it's main timeloop.
+A core library remains, with all numerical algorithms.
+On top of this library a light-weight \ac{API} exposes direct memory access the variables and to the control functions.
 
@@
 More specifically: get current water levels and flow fields, current running state of water pumps, and other structures, active meteorological forcings.
-% Fix sentence
 Each of these state variables can just as easily be \emph{set} as they are \emph{get}.
 These functions are implemented in correspondence with the \ac{BMI} standard \citep{Peckham2013}.
-The \ac{API} exposes memory pointers to relevant model state variables.
 This is how the model directly `feels' and uses its new state.
@@
 
-The freedom of changing model state is possible due to the core and \ac{API} adhering to the Hollywood Principle: ``Don't call us, we'll call you''.
-The core or model does not run itself, but instead is instructed by the outside world.
-Typically, the timeloop from a conventional program is extracted and moved to the calling application; only the single timestep remains in the core.
-
-In several workshops, model coupling frameworks \citep[see][for an overview]{Jagers2010}, have agreed on the minimal interface that makes a model adhere to the Hollywood principles, by exposing the following methods:
+The freedom of interaction is possible due to the core adhering to the Hollywood Principle: ``Don't call us, we'll call you''.
+In several workshops, model coupling frameworks \citep[see][for an overview]{Jagers2010}, have agreed on the minimal interface that makes a model adhere to the Hollywood principles; by exposing the following methods:
 
@@
 
-\section{Architecture}
+\section{Platform architecture}
 Now that a numerical model is available that allows high resolution interactive modeling, how can we hook it up to a tablet, mobile phone and map table, all at the same time, to allow for social and interactive modeling?
@@
 Provisioning, in this context, is preparing a model controller so that it starts its computation on request.
-Machines are prepared and put on hold at one of the \ac{IaaS} providers (such as Amazon).
+The platform runs on GNU/Linux based \ac{IaaS} providers, such as Amazon.
+For each running engine (one customer can have multiple running engines) a machine is reserved.
 These machines are prepared with a container that contains the model schematization, tables and grids, and hot restart files.
@@
 When a user interacts with the model, for example places a discharge point, turns on a pump or lowers an orifice, the user expects a response within a second.
-This short feedback loop is achieved using several techniques.
-User interactions are directly sent through websockets \citep{Hickson2012} to the model controller.
+To achieve this, user interactions are directly sent through websockets \citep{Hickson2012}.
 These interactions are cached and pushed into the model on every timestep.
-These timesteps have to be kept shorter than in a classical model run, \SI{100}{\milli\second} is common.
+Timesteps have to be kept shorter than in a classical model run, \SI{100}{\milli\second} is common.
 After a model timestep, the chain is inverted.
-The model pushes its arrays back in the direction of the browser, which is subscribed to these messages and can directly visualize the results.
-This fast feedback loop supports user interactivity with the model.
+The model pushes its arrays back in the direction of the browser, which is subscribed to these messages and can directly visualize.
 
@@
 The sessions worked best when a topic expert, not necessary a modeler, was present in the session.
+A detailed analysis of how the platform is used in a social context showed that the speed and flexibility determine the suitability \citep{Leskens2014}.
 
@@
 An alternative is to interpolate in time and provide the insinuation of continuous information.
+
 We use both approaches here.
@@
 Several other models use a similar numerical approach, albeit with limitations.
-The subgrid scheme by \citet{Stelling2012} is partly based on the subgrid method for flooding and drying by \citet{Casulli2009}, who implements his methods in the UnTRIM research package \citet{Casulli2000}.
-UnTRIM uses fully unstructured grids as opposed to the quad tree grids used here.
+The subgrid scheme by \citet{Stelling2012} is partly based on the subgrid method for flooding and drying by \citet{Casulli2009}, who implements his methods in the unstructured UnTRIM research package \citet{Casulli2000}.
 The coupled 1D-2D hydrodynamic model MIKE-FLOOD \citep{Dhi2014} also offers subgrid overland flow.
-In the accompanying MIKE-SHE all relevant vertical hydrological processes are offered, also with some degree of subgrid-information for the infiltration part.
 
 % Limitations
-The \threedi platform is suitable for exploratory purposes.
+The high resolution results in great ``face validity''.
+For now, the \threedi platform is suitable for exploratory purposes, as only a few marks of the confidence checklist \citep[][pp.
+136]{Baart2013} have been ticked.
 Work has started on a skillbed for the \threedi platform.
-A skillbed is an automated test suite that compares results to analytical cases, measurements, and results from other numerical models.
+A skillbed is an automated test suite that compares results to a wide range of analytical cases, measurements, and results from other numerical models.
 
@@
 % don't put everything into 1 model
-Another planned step is to incorporate or preferably to couple the system with other physical models.
+Another planned step is to incorporate or preferably to couple the system with other numerical engines.
 The monolithic approach of putting all processes into one model has proven cumbersome in previous implementations.

### 2014-06-27 article.tex
fix bib and full names

@@
 
-\author{\underline{F. Baart}
+\author{\underline{Fedor Baart}
   \address[A1]{\it{Deltares,
@@
       Delft, The Netherlands (fedor.baart@deltares.nl, arthur.vandam@deltares.nl, gennadii.donchyts@deltares.nl)}},
-  K. K. Ha
+  Jack Ha
   \address[B1]{\it{Nelen \& Schuurmans,
@@
       Utrecht, The Netherlands (jack.ha@nelen-schuurmans.nl, martijn.siemerink@nelen-schuurmans.nl)}},
-  A. van Dam \addressmark[A1],
-  G. Donchyts \addressmark[A1],
-  M. Siemerink \addressmark[B1]
+  Arthur van Dam \addressmark[A1],
+  Gennadii Donchyts \addressmark[A1],
+  Martijn Siemerink \addressmark[B1]
 }
